---
title: "DM Econometrics - Labor Market and Origin"
subtitle: "Econometrics 2 — ENSAE 2A — Polixène Gresant, Rachel Issiakou, Juliette Lesné"
date: "Avril 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
#install.packages()
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sandwich)    
library(lmtest)      
library(AER)         
library(plm)         
library(margins)     
library(stargazer)  
library(questionr)
library(estimatr)
library(modelsummary)
```

```{r data, include=FALSE}
url <- "https://github.com/pgresant/econo2_project/raw/main/DM_Subject_2_Data.dta"
df <- read_dta(url)

# We recode the variable to better understand them in the analysis
df$origine <- dplyr::recode(df$origine,   
  "01" = "France",
  "03" = "Nothern Europe",
  "04" = "Southern Europe",
  "05" = "Eastern Europe",
  "06" = "Maghreb",
  "07" = "Rest of Africa",
  "08" = "Middle East",
  "09" = "Southern Asia",
  "10" = "Rest of the world") 
#freq(df$origine)

df$acteu <- dplyr::recode(df$acteu,   
  "1" = "Employed",
  "2" = "Unemployed",
  "3" = "Inactive") 
#freq(df$acteu)

df$csp_actif <- dplyr::recode(df$csp_actif,   
  "1" = "Farmers",
  "2" = "Craftsmen, company head",
  "3" = "CPIS", 
  "4" = "Intermediate profession",
  "5" = "Service employees",
  "6" = "Manual workers", 
  "8" = "Unemployed never worked", 
  "0" = "Not stated"
  ) 
#freq(df$csp_actif)

df <- df %>%
  mutate(
    #contrat
    contrat = na_if(contrat, ""),  # transforms "" into NA
    contrat = dplyr::recode(contrat,
      "0" = "Not employed",
      "1" = "CDI",
      "2" = "CDD",
      "3" = "Seasonal_contract",
      "4" = "Interim",
      "5" = "Apprentissage",
      .missing = "99"), # uses the NA to recode them into 99 
      
    #education
    education = na_if(education, ""),
    education = dplyr::recode(education,   
      "1" = ">Bac+2",
      "3" = "Bac+2",
      "4" = "Bac",
      "5" = "CAP-BEP",
      "6" = "Brevet",
      "7" = "Aucun",
      .missing = "99")
    )
#freq(df$contrat)

df <- df %>% mutate(
  
  origine_bin = ifelse(origine == "France", "France", "Etranger"),
  # We make the assumption that individuals who do not state 
  # their nationalities are treated as non-French
  
  origine_bin99 = case_when(
    origine == "France" ~ "France",
    origine == "99"     ~ "Not Stated",
    TRUE                ~ "Etranger"
  )  # in the binary variable origine_bin99 we do not make the assumption that individuals whom origin is unknown are foreigners. We create a third category.
  
)

df <- df %>% mutate(origine_cat2 = case_when(
  origine == "France" ~ "France",
  origine %in% c("Rest of Africa","Maghreb", "Middle East") ~ "Africa_ME" ,
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat3 = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East",
  origine %in% c("Rest of Africa","Maghreb") ~ "Africa" ,
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat4 = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East", # 30 individuals in lastq_emp
  origine %in% c("Rest of Africa") ~ "Rest of Africa", # 61 individuals in lastq_emp
  origine %in% c("Maghreb") ~ "Maghreb", # 122 individuals in lastq_emp
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East", # 30 individuals in lastq_emp
  origine %in% c("Rest of Africa") ~ "Rest of Africa", # 61 individuals in lastq_emp
  origine %in% c("Maghreb") ~ "Maghreb", # 122 individuals in lastq_emp
  origine %in% c("Southern Europe") ~ "Southern Europe",
  origine %in% c("Eastern Europe") ~ "Eastern Europe",
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))


# We are going to recode some variables in the data set so that we can use them later in our project
class(df$salhoraire) # verification of the class of the dependent variables 
class(df$logsalhoraire)

vars_num <- c("age")
df[vars_num] <- lapply(df[vars_num], as.numeric) # We convert continous variables into numbers, in order to use them in regressions

vars_cat <- c("education", "origine", "contrat", "csp_actif", "csp_mere", "csp_pere", "acteu", "homme", "origine_bin99", "origine_cat", "origine_bin")
df[vars_cat] <- lapply(df[vars_cat], as.factor) # We convert categorical variables to factor, in order to use them in regressions

# We precise the category of reference
df$acteu <- relevel(df$acteu, ref = "Employed")
df$origine <- relevel(df$origine, ref = "France")
df$origine_bin <- relevel(df$origine_bin, ref = "France") # la catégorie aucun est la plus fréquente mais bon ...
df$education <- relevel(df$education, ref = "Aucun")# la catégorie 7 est la plus fréquente mais bon ...je pense on eput changer selon le subdataset
df$contrat <- relevel(df$contrat, ref = "CDI") # category la plus fréquente
df$csp_actif <- relevel(df$csp_actif, ref = "Service employees") # category la plus fréquente$
df$origine_cat <- relevel(df$origine_cat, ref = "France")
df$origine_bin99 <- relevel(df$origine_bin99, ref = "France")
df$origine_bin <- relevel(df$origine_bin, ref = "France")
# We create some variables 
df$age2 <- df$age^2
```

Some descriptive statistics in order to understand the dataset
```{r}
table(df$immi, df$nat)
```
According to INSEE, an immigrant is a person born abroad who is a foreign national and resides in France.
There is 1956 individual*quarter French people that are immigrants. Some immigrants have been able to become French citizens, whilst others remain foreign nationals. Indeed, immigrant status is permanent. 

```{r}
table(df$descimmi, df$origine)
```
People who have at least one parent that held a different nationality than the French nationality can be a descendants of immigrant(s) (their parents are immigrants, meaning that they are born abroad without the French nationality). 
The other part of people who have at least one parent that held a different nationality than the French nationality are not descendant of immigrant (their parent do not held the French nationality but are born in France). 

```{r}
cprop(table(df$immi, df$origine_bin))
```
53,2% of the individualsxquartier who have at least one parent that held a different nationality than the French nationality are immigrant, meaning they are are born abroad without the French nationality. 

# 1. Intensive Margin (Net Hourly Wage)

**Question 1**

```{r q1_sample}
# We select the sample : employed individuals in the last observed quarter (t=6) 
summary(df[df$t == "6" & df$acteu == "Employed", ]$salhoraire) # 660 NA in the dependent variable

lastq_emp <- df %>%
  dplyr::filter(t == 6, acteu == "Employed",  !is.na(salhoraire)) 

cat("N =", nrow(lastq_emp), "employed individuals with observed salary \n")
summary(lastq_emp$logsalhoraire) # no NA in the dependent variable

freq(lastq_emp$annee) # 42% of the observation are made in 2015, 58% in 2016
table(lastq_emp$annee, lastq_emp$trimestre) 
prop(table(lastq_emp$annee, lastq_emp$trimestre)) 

# to take into account the effect of the time gap between the observation, we add an indicator
lastq_emp <- lastq_emp %>% mutate(indic2016 = ifelse(lastq_emp$annee == 2016, 1, 0))
freq(lastq_emp$indic2016) # OK 
```

```{r q1_stats}
# First, we make few descriptive statistics of salary 
summary(last_q$logsalhoraire)

# Distribution of origin within employed individuals
freq(lastq_emp$origine)

# descrptive statistics 
rprop(table(lastq_emp$origine, lastq_emp$immi))
cprop(table(lastq_emp$origine, lastq_emp$immi))

```
In the subset, arounf 75% of the individuals with at least one parent with an African nationality (except Maghrebian) or a Middle East one are immigrant, whereas it's 47% for Magrhebian and Eastern Europe origins, 30% for Northern and Southern European orgins.  

This tendancy is confirmed by the second table : in the immigrant population, there 20% of people with at least one parent with an Maghrebian nationality, 17% with an african nationality and 25% with southern european nationality. 

We estimate several models, using the function lm_robust (instead of lm), to compute heteroscedasticity-robust standard errors. Indee, we are not sure that the homoscedasticity assumptions holds here.

```{r}
m1 <- lm_robust(logsalhoraire ~ indic2016 + origine, data = lastq_emp)
summary(m1) # do not put the * of significance 
modelsummary(m1, stars = TRUE) # allow us to see the * of significance  

m1.1 <- lm_robust(logsalhoraire ~ origine, data = lastq_emp)
summary(m1) ; summary(m1.1) # to compare the 2 models easily 
modelsummary(m1, stars = TRUE) 
```
The estimator of the coefficient of indic2016 is very low (<0,01) and not significant at 10%. The Ajusted R-squared is lower in the odel with the indicator than the other. So, we will not include this variable in the next models. 

In the first regression (m1.1), we regress only the logarithm of wages on origin. The coefficients that are significant at 5% of level of risk are those for the Northern European origin (+0.119), the rest of Africa (-0.177) and South Asia (-0.151). The coefficient of the origin of Middle East origin (+0,09) is significant at the 10% level, but it's less relevant than the others. This can be interpreted as follows: an employee of South Asian origin earns approximately 15% less than an employee of French origin, in average. We can underline that the coefficient of the Maghreb origin has a p-value equal to 10,12%, meaning that it is not significant at 10% level, but not so far from this level. 

Considering the effectifs and the result of the regression, we did the agregation of the variable seen upper in the code. It was necessary to agregated some categories, in order to get effectifs enough large to have be significate, when we will add other variables.
We regrouped : Southern Asia and the Rest of the world ; Easter Europe and Southern Europe, only to see the effect, and to have around 30 indivuals at least in each category. It would be possible after, to agregate the categories Southern Asia and the Rest of the world with the Not Stated (99), in terms of effect. 

```{r}
m1.2 <- lm_robust(logsalhoraire ~ origine_cat, data = lastq_emp)
summary(m1.2)
freq(lastq_emp$origine_cat)
```
These results where expected.  

```{r}
m1.3 <- lm_robust(logsalhoraire ~ origine_bin, data = lastq_emp)
summary(m1.3)
```
The dummy variable origin is significant at the 5% level. An employee of that have at least on parent who do not held the french Nationality earns approximately 3% less than an employee of French origin, in average and without variable of control. This can be interesting for us, in order to have an aggregated of this social condition, but it is not precise enough. 


```{r}
m2 <- lm_robust(logsalhoraire ~ immi + origine, data = lastq_emp)
m2.1 <- lm_robust(logsalhoraire ~ immi + origine_bin, data = lastq_emp)
m2.2 <- lm_robust(logsalhoraire ~ immi + origine_cat, data = lastq_emp)
summary(m2); summary(m2.1) ; summary(m2.2)
```
When we control by the dummy variable "immi", the origin is not . 
- In model 2.1, the variable origin is not statistically significant.  
- In model 2.2, the variable of the African origin is not significant anymore. The only significant variable are the open of Nothern European origin and Not Stated, that are more significant than before (at level 1%), and the coefficient are more important for the Northern origin (+3 point of pourcentage), meaning that the effect of this origin is broader when controlling by the immigration status. The variable of the middle East origin is significant at 10%.
- In model 2, 

Pb d'endogéniété car plusieurs corrélation entre les régresseurs. on risque ici d'ajouter un biais de variable incluse. Donc on le rajoute pa la variable immi en controle. Néanmoins, une variable d'interaction pourrait etre pertinente. 
=> ou alors on parle de cannal ntm pr les personnes africa origin??

We add an interaction variable in the model. 
```{r}
m2.3 <- lm_robust(logsalhoraire ~ immi*origine_cat, data = lastq_emp)
summary(m2.3)
```
Rien est significatif. pb effectif et biais variable incluse potentiel. on garde pas cette specification qui n'est pas pertinente. 

```{r}
m3 <- lm_robust(logsalhoraire ~ origine_cat + homme + age + age2 + contrat, data = lastq_emp)
summary(m3)
```
In this model, we regress only the logarithm of wages against gender, age, age squared, and origin. 
In absolute value, the coefficients : 
- increase for Middle East origin and Northern Europe origin variables.
- decrease for the Rest of Africa and Rest of the world origin variables.  

En termes de significativité statistique : -
- la variable d'origine du Middle East devient statistiquement significative à 10%, 
- la variable d'origine Rest of africa reste statistiquement significative à 5% (p-value of 1,06%). 
- la variable d'origine d'Europe du Nord est toujours statistiquement significative à 1%. 

```{r}
m4 <- lm_robust(logsalhoraire ~ origine + homme + age + age2 + csp_actif, data = lastq_emp)
summary(m4)
```

In this model, we regress only the logarithm of wages against gender, age, age squared, and origin. 
In absolute value, the coefficients : 
- decrease for Middle East origin, Northern Europe, and Rest of the world origin variables. origin variables.
- increase for the Rest of Africa   

En termes de significativité statistique : -
- la variable d'origine du Middle East is not statiscally significant at 10%, 
- la variable d'origine Rest of africa reste statistiquement significative à 5%. 
- la variable d'origine d'Europe du Nord reste statistiquement significative à 5%.  

The R-squared is much higjer in this model than the precedent one. 
```{r}
m5 <- lm_robust(logsalhoraire ~ homme + age + age2 + origine_cat + contrat + csp_actif, data = lastq_emp)
summary(m5)
```

Les effectifs ne sont pas assez grands pour avoir autant de variables de contrôle et conserver une inférence robuste.  

```{r}
m6 <- lm_robust(logsalhoraire ~ origine + education + homme + age + age2 + csp_actif, data = lastq_emp)
summary(m6)
```
correlation entre education et entre csp_actif (mais très agrégé)

Comparing to the model 4, in this model : 

In absolute value, the coefficients : 
- decrease for Rest of Africa, Northern Europe, Rest of the worl and Not stated origin variables. 

En termes de significativité statistique : -
- la variable d'origine du Middle East is not statiscally significant at 10%,
- la variable d'origine Rest of africa reste statistiquement significative à 10%. 
- la variable d'origine d'Europe du Nord reste statistiquement significative à 5%.  
- the variable of origin Southern Europe significant à 5%, which was not the case in the previous model. 

```{r}
m7 <- lm_robust(salhoraire ~ homme + age + age2 + origine + csp_actif, data = lastq_emp)
summary(m7)
```

**Explications**

Four different regressions are carried out to test the differences in the coefficients of the variables and their significance, depending on the variables and categories used. These are four OLS models with standard errors robust to heteroscedasticity (HC1) based on the 4 046 employees in the last quarter. In the first three regressions, the dependent variable is the log of the hourly wage; whereas in the last one, the dependent variable is the hourly wage in absolute terms. For the first three regressions, explanatory variables are added in each case to determine the channels through which origin affects wages.

In the first regression, we regress only the logarithm of wages against gender, age, age squared, and origin. The significant coefficients relating to origin are those for Northern European origin (+0.126), the rest of Africa (-0.15) and South Asia (-0.151). This can be interpreted as follows: all other things being equal, an employee of South Asian origin earns approximately 15% less than an employee of French origin.

In the second regression, the education variable is included. The origins with significant coefficients are those from Southern Europe (0.49) and the rest of Africa (-0.113). The origin coefficients decrease for almost all origins. For example, for the ‘rest of Africa’ origin, the wage gap relative to individuals of French origin is -15% in the first regression, then -11% in the second. This means that part of this wage gap is due to the fact that certain groups have, on average, a lower level of education. For those of South-East Asian origin, the wage gap with employees of French origin disappears when education is included, which therefore appears to fully explain the gross wage gap.

In the third regression, explanatory variables relating to contract type and socio-professional categories are also included. With regard to origin, the significant coefficients are those for origins associated with Northern Europe (+0.1) and Southern Europe (0.0577). The coefficients for origin diminish further and become insignificant. So another part of the pay gap can now be explained by the type of job held. The coefficient associated with African origin is no longer significant, so does the pay gap relative to individuals of French origin disappear completely once we control for the variables of education and type of job. Thus, direct discrimination regarding equal pay for equal work is not captured here, as inequality operates upstream, via access to education and good jobs. Nevertheless, there may be over-control if these variables are themselves consequences of origin.

For the final regression, we take wage as the dependent variable directly, and as a result the coefficients associated with the intercept are no longer significant. This can be explained by the high variance in wage levels, as the distribution is skewed and high wages introduce noise. In the following analysis, we will therefore use the logarithm of wage as the dependent variable;

In conclusion, wage gaps gradually narrow as we add controls. The only group for which the gap persists after controls is Southern Europe (with a coefficient of +6% in the third regression). So, at first glance, origin acts mainly through observable channels (level of education, type of employment), rather than through direct discrimination when the job and qualifications are identical.


**Question 2**
*Question 2a*
Education is a potentially endogeneous variable in the regression. Indeed, the level of education of an individual partly results from his/her own choices, which are influenced by unobservable charateristics (such as family cultural capital, ability, motivation, ...). Theses characteristics also directly affect wages. Then, we expect that the correlation between education and the residuals is not null, which violates the exogeneity assumption required for OLS to be consistent. Including education in the model therefore introduces an endogeneity bias, and then, the OLS estimator of the education coefficient does not converge to the true causal effect. 

Furthermore, origin and education are correlated. Individuals from certain origin groups have on average lower educational attainment, due to inequalities in access to education. Then, part of the origin on wages operates through the education channel (origin affects education, which in turn affects wages). 
By comparing the short and long regression, including or not the education, we have : 

- when education is omitted in the regression , the origin coefficient captures both the direct effect of origin and its indirect effect through education.
- when education is included in the regression, the origin coefficient is interpreted as the effect of origin conditional on education level. But, as education is endogeneous, the conditional effects is itself biased. In addition, by conditioning on education, we are absorbing part of the total effect of origin on wages : the estimated origin coefficient therefore measures only the direct effect of origin on wages, net of education, and underestimates the total wage inequality due to origin. 

By the omitted variable bias formula (Frisch-Waught):

$$\tilde{\beta}_{origine} = \beta_{origine} + \beta_{educ} \cdot \delta$$

where $\tilde{\beta}_{origine}$ is the coefficient of origin in the short regression of wages on origin ; $\beta_{origine}$ is the coefficient of origin in the long regression of wages on origin and education ; and $\delta = \frac{Cov(educ, origine)}{Var(origine)}$ is the coefficient of the regression of education on origin. As $\beta_{educ} > 0$ and $\delta < 0$ for disadvantaged groups, the bias is negative : we over-estimate in absolute value the wage's disadvantage of theses groups when we omit education. In this case, we attribute the disadvantage for the origins, where it is partly because theses individuals have less education levels, in mean. When we control by education, disadvantage due to origin decreases. 

*Question 2b*

We can choose the variables csp_pere and csp_mere (socio-professional category of the father and mother) as instrumental variables for education. We the construct four binary instruments by regrouping the categories along the PCS nomenclature : 

- father_manager : the father is a senior manager or professional (codes 31 (liberal professions), 33 (senior civil servants), 34 (professors and scientific professions), 35 (information, arts, and entertainment professions), 37 (senior administrative and commercial managers in companies), 38 (engineers and senior technical managers in companies))
- mother_manager : the mother is a senior manager or professional (same codes)
- father_worker : the father is a manual worker (codes 62 (skilled industrial workers), 63 (skilled craft workers), 64 (drivers), 65 (skilled handling, ware-housing, and transport workers), 67 (unskilled industrial workers), 68 (unskilled craft workers), 69 (agricultural workers))
- mother_worker : the mother is a manual worker (same codes)

```{r}
manager   <- c(31, 33, 34, 35, 37, 38)
worker <- c(62, 63, 64, 65, 67, 68, 69)

lastq_emp <- lastq_emp %>%
  mutate(
    csp_pere_num = suppressWarnings(as.numeric(as.character(csp_pere))), # the values are converted to characters and then to numbers
    csp_mere_num = suppressWarnings(as.numeric(as.character(csp_mere))),
    father_manager   = as.numeric(csp_pere_num %in% manager),
    mother_manager   = as.numeric(csp_mere_num %in% manager),
    father_worker = as.numeric(csp_pere_num %in% worker),
    mother_worker = as.numeric(csp_mere_num %in% worker)
  ) %>%
  filter(!is.na(father_manager), !is.na(mother_manager)) # We only keep the rows where "father_manager" is not missing and "mother_manager" is not missing.
# other socio-occupational categories are the default category

# We convert education variable into numeric variable (education_num) to make the 2SLS (education_num is then continuous and we have just one endogeneous variable to instrument, that is possible with 4 instruments)
lastq_emp <- lastq_emp %>%
  mutate(
    education_num = case_when(
      education == ">Bac+2"  ~ 1,
      education == "Bac+2"   ~ 2,
      education == "Bac"     ~ 3,
      education == "CAP-BEP" ~ 4,
      education == "Brevet"  ~ 5,
      education == "Aucun"   ~ 6,
      TRUE                   ~ NA_real_
    )
  )
```
  

Discussion of the validity of the instruments : 2 conditions for an instrumental variable : 

1) Relevance condition (i.e. cov(instrument, education) not equal to 0)
In the sociological and economic literature, we know that parental socio-professional category is highly correlated with the educational attainment of children. Children of senior managers and professionals have substantially higher probabilities of attaining higher education, while children of manual workers have substantially lower probabilities. 
Emprically, we will confirm that by the first-stage F-statistic

2) Exogeneity condition (i.e. cov (instrument, residuals) = 0)
This condition requires that parental ctageory socio-professionnal affects the individual's wage only through the education channel, and not directly.

The respect of this condition is more debatable, because it remains social capital transmission : for example, children of senior managers may benefit from parental networks, job referrals, or behavioural norms that directly improve their labour market outcomes independently of the diploma they hold. In that case, the instrument is invalid.

To identify a model with instrumental variables, the number of instruments must be superior or equal to the number of endogeneous variables. Here, we have only one variable endogeneous (education), and 4 instruments. That improves first-stage precision and allows us to test the joint validity of the instruments via the Sargan-Hansen test. The null hypothesis of this test is that all instruments are valid. 
```{r}
summary(iv_model, diagnostics = TRUE)
```
*Question 2c*

First stage
```{r}
first_stage <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif +
    father_manager + mother_manager + 
    father_worker  + mother_worker,
  se_type = "stata",
  data = lastq_emp
)
summary(first_stage)
```

We examine the coefficients of the four instruments individually in the first-stage regression.
All four coefficients are statistically significant at the 1% level : father_manager (−0.510, p < 0.001), mother_manager (−0.382, p < 0.001), father_worker (+0.233, p < 0.001) and mother_worker (+0.179, p < 0.01).

The signs are consistent with economic intuition : having a parent who is a senior manager or professional is associated with a higher level of education (lower education_num); while having a parent who is a worker is associated with a lower level of education (higher education_num).

To confirm this formally, we conduct a joint significance test of the four instruments.
```{r}
# We use the Wald test with : 
# H0 : the coefficients for all four instruments are jointly zero (irrelevant instruments)
# vs H1 : at least one instrument is relevant
# Simplified model (without instruments)
first_stage_restricted <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif,
  se_type = "stata",
  data = lastq_emp
)

waldtest(first_stage, first_stage_restricted)

F_stat <- 201.3 / 4  # Chi² divided by the number of restrictions
cat("F-stat :", round(F_stat, 2), "\n")

```

The Wald test yields Chi²(4) = 201.3 (p < 0.001), which is equivalent to F ≈ 50.33, far above the conventional threshold of 10. We therefore reject the null hypothesis that the instruments are jointly irrelevant, which confirms that parental socio-professional category is strongly correlated with the individual's education level.


Now, we make the 2SLS :
```{r}
iv_model <- iv_robust(
  logsalhoraire ~ homme + age + age2 + origine + csp_actif + education_num |
                  homme + age + age2 + origine + csp_actif +
                  father_manager + mother_manager +
                  father_worker  + mother_worker,
  se_type = "stata",
  data = lastq_emp
)
summary(iv_model, diagnostics = TRUE)
```
The coefficient of education_num is significative at 1% level and negative, which is consistent with the expected relationship between education and wages (as the level of education increases, so does the wage)
# 2. Extensive Margin (Employed vs. Unemployed)

```{r}

```

**Question 6**

Question 6a : 

Question 6b : 

```{r}
# subset 
lastq <- df %>%
  dplyr::filter(t == 6, acteu != "Inactive") 
cat("N =", nrow(lastq), "active individuals")

# recoding for modeling 
lastq <- lastq %>% mutate(act = ifelse(lastq$acteu == "Employed", 1, 0))

# Checking
#freq(lastq$act)
#freq(lastq$acteu)
```

We begin by estimating a probit model. 
```{r}
probit_model <- glm(act ~ origine + homme + contrat + pcs + age, family = binomial(link = "probit"), data = lastq)
summary(probit_model) # we interpret the sign and the significance of the coefficients, but not their values 

# Average Marginal Effects
AME_probit <- margins(probit_model)
summary(AME_probit)

```

We estimate a logit model. 
```{r}
logit_model <- glm(act ~ origine, family = binomial(link = "logit"), data = lastq)
summary(logit_model) # we interpret the sign and the significance of the coefficients, but not their values 

# Average Marginal Effects
AME_logit <- margins(logit_model)
summary(AME_logit)
```

Question 6c : 

We create the indicator of having a higher education degree (modality "1" of the variable education in the dataset). 
```{r}
lastq <- lastq %>% mutate(highdegree = ifelse(lastq$education == ">Bac+2", 1, 0))
# Checking
#freq(lastq$highdegree) # 975 individuals
#freq(lastq$education)
```



partie de base à garder pr avoir une idée de comment faire ce type de truc  
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
