---
title: "DM Econometrics - Labor Market and Origin"
subtitle: "Econometrics 2 — ENSAE 2A — Polixène Grésant, Rachel Issiakou, Juliette Lesné"
date: "Avril 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

```{r setup}
#install.packages()
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sandwich)    
library(lmtest)      
library(AER)         
library(plm)         
library(margins)     
library(stargazer)  
library(questionr)
library(estimatr)
library(modelsummary)
```

```{r data}
url <- "https://github.com/pgresant/econo2_project/raw/main/DM_Subject_2_Data.dta"
df <- read_dta(url)

# We recode the variable to better understand them in the analysis
df$origine <- dplyr::recode(df$origine,   
  "01" = "France",
  "03" = "Nothern Europe",
  "04" = "Southern Europe",
  "05" = "Eastern Europe",
  "06" = "Maghreb",
  "07" = "Rest of Africa",
  "08" = "Middle East",
  "09" = "Southern Asia",
  "10" = "Rest of the world") 
#freq(df$origine)

df$acteu <- dplyr::recode(df$acteu,   
  "1" = "Employed",
  "2" = "Unemployed",
  "3" = "Inactive") 
#freq(df$acteu)

df$csp_actif <- dplyr::recode(df$csp_actif,   
  "1" = "Farmers",
  "2" = "Craftsmen, company head",
  "3" = "CPIS", 
  "4" = "Intermediate profession",
  "5" = "Service employees",
  "6" = "Manual workers", 
  "8" = "Unemployed never worked", 
  "0" = "Not stated"
  ) 
#freq(df$csp_actif)

df <- df %>%
  mutate(
    #contrat
    contrat = na_if(contrat, ""),  # transforms "" into NA
    contrat = dplyr::recode(contrat,
      "0" = "Not employed",
      "1" = "CDI",
      "2" = "CDD",
      "3" = "Seasonal_contract",
      "4" = "Interim",
      "5" = "Apprentissage",
      .missing = "99"), # uses the NA to recode them into 99 
      
    #education
    education = na_if(education, ""),
    education = dplyr::recode(education,   
      "1" = ">Bac+2",
      "3" = "Bac+2",
      "4" = "Bac",
      "5" = "CAP-BEP",
      "6" = "Brevet",
      "7" = "Aucun",
      .missing = "99")
    )
#freq(df$contrat)

df <- df %>% mutate(
  
  origine_bin = ifelse(origine == "France", 0, 1),
  # We make the assumption that individuals who do not state their nationalities are treated as non-French
  
  origine_bin99 = case_when(
    origine == "France" ~ "France",
    origine == "99"     ~ "Not Stated",
    TRUE                ~ "Etranger"
  )  # In the binary variable origine_bin99, we don't assume that individuals whom origin is unknown are foreigners. We create a third category.
  
)

df <- df %>% mutate(origine_cat2 = case_when(
  origine == "France" ~ "France",
  origine %in% c("Rest of Africa","Maghreb", "Middle East") ~ "Africa_ME" ,
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat3 = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East",
  origine %in% c("Rest of Africa","Maghreb") ~ "Africa" ,
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat4 = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East", # 30 individuals in lastq_emp
  origine %in% c("Rest of Africa") ~ "Rest of Africa", # 61 individuals in lastq_emp
  origine %in% c("Maghreb") ~ "Maghreb", # 122 individuals in lastq_emp
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East", # 30 individuals in lastq_emp
  origine %in% c("Rest of Africa") ~ "Rest of Africa", # 61 individuals in lastq_emp
  origine %in% c("Maghreb") ~ "Maghreb", # 122 individuals in lastq_emp
  origine %in% c("Southern Europe") ~ "Southern Europe",
  origine %in% c("Eastern Europe") ~ "Eastern Europe",
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))


# We are going to recode some variables in the data set so that we can use them later in our project
class(df$salhoraire) # verification of the class of the dependent variables 
class(df$logsalhoraire)

vars_num <- c("age", "origine_bin")
df[vars_num] <- lapply(df[vars_num], as.numeric) # We convert continuous variables into numbers, in order to use them in regressions

vars_cat <- c("education", "origine", "contrat", "csp_actif", "csp_mere", "csp_pere", "acteu", "homme", "origine_bin99", "origine_cat")
df[vars_cat] <- lapply(df[vars_cat], as.factor) # We convert categorical variables to factor, in order to use them in regressions

# We precise the category of reference
df$acteu <- relevel(df$acteu, ref = "Employed")
df$education <- relevel(df$education, ref = "Bac") # we choose an intermediate category 
df$contrat <- relevel(df$contrat, ref = "CDI") # most frequent category
df$csp_actif <- relevel(df$csp_actif, ref = "Service employees") # most frequent category
df$origine_cat <- relevel(df$origine_cat, ref = "France")
df$origine_bin99 <- relevel(df$origine_bin99, ref = "France")

# We create the variable age squared
df$age2 <- df$age^2
```

For the purposes of our analysis, we have chosen to include ‘not stated’ responses of the variables, so as not to skew the sample.

Fist, we compute some descriptive statistics in order to understand the dataset and to precisely define being an immigrant and/ or a foreigner in France. It's important to have in mind this definition for the treatments that follow, specifically to understand discrimination in the labor market. 
```{r}
table(df$immi, df$nat)
```
According to INSEE, an immigrant is a person born abroad without the French nationality who resides in France.
There are 1956 individual*quarter French people that are immigrants. Some immigrants are French citizens, whilst others remain foreign nationals. The immigrant status is permanent. 

```{r}
table(df$descimmi, df$origine)
```
Individuals born in France with at least one parent of foreign nationality are classified as descendants of immigrants. 
The remaining individuals with at least one parent of a different nationality are not necessarily descendants of immigrants (their parent do not hold the French nationality but are born in France). 

```{r}
cprop(table(df$immi, df$origine_bin))
```
53,2% of the individuals*quarter who have at least one parent that holds a different nationality than the French nationality are immigrants, meaning they are born abroad without the French nationality.

We provide the correct version of the variable salhoraire. As this does not affect the results when using the logarithm as the dependent variable (except for the constant/intercept) and as we did the treatment before the correction, we will keep the original variable logsalaire.
```{r}
df$salhoraire = df$salmensuel / (df$heuretra * 4.33)
```


# 1. Intensive Margin (Net Hourly Wage)

## Question 1

### Data and descriptive statistics to understand the issues

```{r q1_sample}
# We select the sample : employed individuals in the last observed quarter (t=6) 
summary(df[df$t == "6" & df$acteu == "Employed", ]$salhoraire) # 660 NA in the dependent variable

lastq_emp <- df %>%
  dplyr::filter(t == 6, acteu == "Employed",  !is.na(salhoraire)) 

cat("N =", nrow(lastq_emp), "employed individuals with observed salary \n")
summary(lastq_emp$logsalhoraire) # no NA in the dependent variable ; descriptive statistics 

freq(lastq_emp$annee) # 42% of the observations are made in 2015, 58% in 2016
table(lastq_emp$annee, lastq_emp$trimestre) 
prop(table(lastq_emp$annee, lastq_emp$trimestre)) 

# to take into account the effect of the time gap between the observation, we add an indicator
lastq_emp <- lastq_emp %>% mutate(indic2016 = ifelse(lastq_emp$annee == 2016, 1, 0))
freq(lastq_emp$indic2016) # check : OK  
```

First, we provide descriptive statistics for both the dependent variable (salary) and the independent variable (origin) to understand the heterogeneity of the data.
```{r q1_stats}
# summary(last_q$logsalhoraire)

# Distribution of origin within employed individuals
freq(lastq_emp$origine)

# descriptive statistics 
rprop(table(lastq_emp$origine, lastq_emp$immi))
cprop(table(lastq_emp$origine, lastq_emp$immi)) ; freq(lastq_emp$origine) # in order to compare
```
In the subset, around 75% of the individuals with at least one parent with an African nationality (except North African origin) or a Middle East nationality are immigrants, whereas it's only 47% for North African and Eastern Europe origins and 30% for Northern and Southern European origins. These individual don't have the same status, and can face different types of discrimination in the labor market.   

This tendency is confirmed by the second table : in the immigrant population, there are 20% of people with at least one parent with a North African nationality, 17% with an African nationality and 25% with Southern European nationality. 

### The choice of the dependent variable 

We estimate several models, using the function lm_robust (instead of lm), to compute heteroscedasticity-robust standard errors. We are not sure that the homoscedasticity assumptions holds here. We begin by regressing the logarithm of the hourly wage on the origin and the indicatrice of the year.  

```{r}
m1 <- lm_robust(logsalhoraire ~ indic2016 + origine, data = lastq_emp)
summary(m1) # does not put the * of significance

m1.1 <- lm_robust(logsalhoraire ~ origine, data = lastq_emp)
summary(m1) ; summary(m1.1) # to compare the 2 models easily 
```
The estimator of the coefficient of indic2016 is very low (<0,01) and not significant at 10%. The Ajusted R-squared is lower in the model with the indicator than the other. So, we will not include this variable in the next models. 

In the first regression (m1.1), we regress only the logarithm of wages on origin. The coefficients that are significant at 5% of level of risk are those for the Northern European origin (+0.12), the rest of Africa (-0.18) and South Asia (-0.18). This can be interpreted as follows : a worker of South Asian origin earns approximately 18% less than a worker of French origin, in average. 

Considering the numbers of people in the dataset and the results of the regression, we aggregated the variables defined above in the code. It was necessary to aggregate some categories, in order to ensure that each category has a large enough sample size to yield significant results, when we will add other covariates.

We regrouped Southern Asia and the Rest of the world ; Eastern Europe and Southern Europe, only to see the effect, and to have around 30 or 60 individuals at least in each category. We also added another specification without the aggregation of the Easter Europe and Southern Europe. 

```{r}
m1.2 <- lm_robust(logsalhoraire ~ origine_cat4, data = lastq_emp)
summary(m1.2)
freq(lastq_emp$origine_cat4)

m1.22 <- lm_robust(logsalhoraire ~ origine_cat, data = lastq_emp)
summary(m1.22)
freq(lastq_emp$origine_cat)
```
These results were expected. 

We test the binary variable origin. 
```{r}
m1.3 <- lm_robust(logsalhoraire ~ origine_bin, data = lastq_emp)
summary(m1.3)
```
The dummy variable origin is significant at the 5% level. A worker that has at least one parent who is not French earns approximately 3% less than a worker of French origin, in average and without variable of control. This can be interesting for us, in order to obtain an aggregate of these social conditions, but it is not precise enough, and the risk of mixing heterogeneous situation is very high. So, we will first not use this covariate. 

We test the link with the status of immigrant, as we saw before the difference according to the origin. 
```{r}
m2 <- lm_robust(logsalhoraire ~ immi + origine, data = lastq_emp)
m2.1 <- lm_robust(logsalhoraire ~ immi + origine_bin, data = lastq_emp)
m2.2 <- lm_robust(logsalhoraire ~ immi + origine_cat, data = lastq_emp)
summary(m2); summary(m2.1) ; summary(m2.2)
```
When we control by the dummy variable of being an immigrant, we found these result :  
- In model 2.1, the binary variable origin is not statistically significant.  
- In model 2.2, the same dummy variables of origins are significant : Africa, Northern Europe and Not Stated. 

Northern European origin and Not Stated are more significant than before (at level 1%), but it is the opposite situation for the "Rest of Africa" origin. The coefficients are more important for the Northern origin (+3 points of percentage) meaning that the effect of this origin is broader when controlling by the immigration status. We can think that it is because people with a Northern European origin are less likely to be immigrants, and being an immigrant is negatively correlated to the salary. The coefficient is more important for the "Rest of Africa" origin, meaning that the effect of this origin is less important in absolute value when controlling by the immigration status. We can think that it is because people with an African origin are more likely to be immigrants. 

The immigrant status first seems to be one of the channels of the effect of origin. But, the simultaneous inclusion of both origin and migration status leads to significant collinearity, which undermines the reliability of the estimated coefficients. There is probably a included variable bias, controlling by the immigrant status. Immigrant status is linked to certain geographical trajectories in our sample. Moreover, this choice is justified by the desire to estimate the total effect of origin on wages without controlling for mechanisms linked to the migration transition. 

Due to this conclusion and choice of analysis, we won't present the model with an interaction between 'immi' and 'origin'. 

To conclude our analysis, in this part of question 1, we try to understand the effect of the variable origin, though different canals, leading to different specification of the variable itself. First, in order to apply the Central Limit Theorem and to have a consistent estimator, it is a necessity to have at least 30 individuals in each category. To do this, we grouped certain categories within the variable. We performed various groupings to identify the most relevant one — that is, the grouping that combines the most closely related categories and allows us to identify potential differences in wages (or employment) based on origin. We therefore present only a selection of the specifications tested during this exercise. We initially adopted preliminary groupings to ensure larger sample sizes $>60$) for more robust inference. However, when we examined more detailed variable specifications (with sample sizes $>30$), significant differences between groups became apparent, especially during the analyse of the unemployement. Consequently, we chose to retain the most detailed variable possible and the same variable for both part of the work to coherent. 

### Specification with other covariates

We are interesting by controlling our model by covariates to see some differences of wage that are not justify, and considered as unfair. First, we compute add the socio-demographics covariates and the type of contract in the linear model. 
```{r}
m3 <- lm_robust(logsalhoraire ~ origine_cat + homme + age + age2 + contrat, data = lastq_emp)
summary(m3)
```
In this model, we regress the logarithm of wages on gender, age, age squared, origin and the type of contract. 

In terms of statistical significance : 
- the variable Middle East origin becomes statistically significant at the 10% level, 
- the original variable ‘Rest of Africa’ remains statistically significant at the 5% level. 
- the original variable for Northern Europe and “Not Stated” remains statistically significant at the 1% level.
- all the dummy contract variable are statistically significant at 1%. 

Qualitatively, comparatively to being in a long term contract (CDI), the fact of being in more precarious situations (CDD, Interim, Seasonal contract, Apprentissage) is associated with a decrease of the hourly wage, on average.   

Concerning variable of interest and from a quantitative point of view, all else being equal, having one parent with an African nationality (except one from Maghreb) is associated with a 12% reduction of the hourly wage, in average. However,  all else being equal, having one parent with a Northern European nationality is associated with a 14% increase of the hourly wage, in average. 

Compared to the previous model m1.22, without variables of control, in absolute value, the coefficients : 
- increase for Middle East origin and Northern Europe origin variables,
- decrease for the Rest of Africa origin. 

The type of contract is therefore one of the channels through which the total effect of the origin is transmitted. The individuals with an African origin tend more likely to have a more precarious contract, unlike the individuals with a Northern European origin. So, controlling by the type of contract allows us to see the net effect of the origin variable. Nevertheless, this structural effect can also be capture by the social position and the type of profession. Indeed, the differences of salaries between professions are very high. In consequence, we test another model with control variable and the csp of the individual, without the type of contract in a first time. 


```{r}
m4 <- lm_robust(logsalhoraire ~ origine_cat + homme + age + age2 + csp_actif, data = lastq_emp)
summary(m4)
```
In this model, we regress the logarithm of wages on gender, age, age squared, origin and csp_actif 

In terms of statistical significance : 
- the variable Middle East origin is not statistically significant, 
- the original variables ‘Rest of Africa’ and Northern Europe remains statistically significant at the 5% level, but not at 1%. 
- the dummy csp variable of CPIS and Intermediate profession are statistically significant at 1%, whereas the other CSP are not significant. It is not a surprise, because the pay conditions of employees and manual worker are very similar nowadays. 

Qualitatively, comparatively to being an employee, the fact of being in a CPIS or in a intermediate profession is associated with a increase of the hourly wage, on average.   

Concerning or variable of interest and from a quantitative point of view, all else being equal, having one parent with an African nationality (except one from Maghreb) is associated with a 7% reduction of the hourly wage, in average. However,  all else being equal, having one parent with a Northern European nationality is associated with a 10% increase of the hourly wage, in average. 

Compared to the previous model m1.22, without variables of control, in absolute value, the coefficients are less important. The wage differences are lower.  

The csp is therefore one of the channels through which the total effect of the origin is transmitted. The individuals with an African origin tend to have a less well-paid csp, unlike the individuals with a Northern European origin. So, controlling by the csp allows us to see the net effect of the origin variable. It is an important variable of contract. 

The adjusted R-squared is much higher in the model including the csp of the individuals than in the precedent one, which include the type of contracts. The model with the csp explains 31% of the variance of our dependent variable after adjusting for the number of variables included, whereas the one with the type of contract explains only 10%, after adjusting for the number of variables included. The variable csp_actif has a better power of explanation in our model.

We test the model with both of these variables. 
```{r}
m5 <- lm_robust(logsalhoraire ~ homme + age + age2 + origine_cat + contrat + csp_actif, data = lastq_emp)
summary(m5)
```
If we include the contract type, we are measuring the effect of the origin background 'at the same contract type' and 'at the same csp'. This model only explains 3 additional points of percentage in the proportion of variance explained. The coefficient of the "Rest of Africa" origin and of the Maghreb origin are not significant anymore, whereas the one of the Northern Europe is. In a surprising way, the coefficient of the Southern Europe origin is significant at 10%. The p-value of this coefficient was much higher in the other models. Since, the coefficient is only significant à 10%, we cannot be sure of the robustness of this model. Moreover, we thin that the sample sizes are not large enough to include so many control variables and still maintain robust inference. We rapidly check the numbers : 

```{r}
table(lastq_emp$contrat, lastq_emp$origine )
```
There are really few people that have not a French origin and another type of contract than CDI. 

Wwe chose to keep the csp instead of the contract type, because we chose to keep the better model with the higher adjusted R-squared and consider the trade-off between the power of explanation and the problem of number of people in all categories. The more explanatory model reduces the correlation between the residuals and the origin variable, thereby providing a clearer understanding of the effect of origin as a whole. Furthermore, the CSP is a status that is less likely to change over the course of an individual’s life than the contractual variable, which is more time-bound and linked to a specific stage of a person’s working life (apprenticeship, fixed-term contract at the start of a career).

In another hand, the effect of education on wage are really important, since it determins productivity and We test the chosen specification, adding the education as a covariate. 
```{r}
m6 <- lm_robust(logsalhoraire ~ origine_cat + homme + age + age2 + csp_actif + education, data = lastq_emp)
summary(m6)
```
In this model, we regress the logarithm of wages on gender, age, age squared, origin, csp_actif and education. 

In terms of statistical significance : 
- the variable Middle East origin is not statistically significant, 
- the original variable ‘Rest of Africa’ and Northern Europe remains statistically significant respectivelly at the 10% level and at 5%. 
- all the dummy education variable of CPIS and Intermediate profession are statistically significant at 1%. It is not a surprise. 

Qualitatively, comparatively to being an employee, the fact of having a higher degree than the Baccalaureat is associated with an increase of the hourly wage, on average, whereas having a lower degree than the Baccalaureat is associated with a decrease of the hourly wage, on average.   

Concerning or variable of interest and from a quantitative point of view, all else being equal, having one parent with an African nationality (except of Maghreb) is associated with a 7% reduction of the hourly wage, in average. However, all else being equal, having one parent with a Northern European nationality is associated with a 9% increase of the hourly wage, in average. 

Compared to the previous model, without variables of control, the coefficients are very similar.

The education is therefore one of the channels through which the total effect of the origin is transmitted. But, it imply endogenous problem, tackled latter.  


Finally, we test to regress the salary on the covariates chosen. 
```{r}
m7 <- lm_robust(salhoraire ~ homme + age + age2 + origine_cat + csp_actif, data = lastq_emp)
summary(m7)
```
The results are quite different than the other models. The origin variables are not significant except for Northern Europe, which is significant at only 10%. So, we will not use this specification afterwards. Moreover, we have chosen to use the logarithm of wages rather than wages themselves, as this makes the results easier to interpret and reduces the dispersion of wages, which is highly asymmetric. 


## Question 2
### Question 2a

Education is a potentially endogeneous variable in the regression. Indeed, the level of education of an individual partly results from his/her own choices, which are influenced by unobservable characteristics (such as family cultural capital, ability, motivation, ...). These characteristics also directly affect wages. We expect that the correlation between education and the residuals is not null, which violates the exogeneity assumption required for OLS to be consistent. Including education in the model therefore introduces an endogeneity bias, and then, the OLS estimator of the education coefficient does not converge to the true causal effect. 

Furthermore, origin and education are correlated. Individuals from certain origin groups have on average lower educational attainment, due to inequalities in access to education. Then, part of the origin on wages operates through the education channel (origin affects education, which in turn affects wages). 
By comparing the short and long regression, including or not the education, we have : 

- when education is omitted in the regression, the origin coefficient captures both the direct effect of origin and its indirect effect through education.
- when education is included in the regression, the origin coefficient is interpreted as the effect of origin conditional on education level. But, as education is endogeneous, the conditional effects is itself biased. In addition, by conditioning on education, we are absorbing part of the total effect of origin on wages : the estimated origin coefficient therefore measures only the direct effect of origin on wages, net of education, and underestimates the total wage inequality due to origin. 

By the omitted variable bias formula (Frisch-Waught):

$$\tilde{\beta}_{origine} = \beta_{origine} + \beta_{educ} \cdot \delta$$

where :
$$\tilde{\beta}_{origine}$$ is the coefficient of origin in the short regression of wages on origin ; 
$$\beta_{origine}$$ is the coefficient of origin in the long regression of wages on origin and education ; 
$$\delta = \frac{Cov(educ, origine)}{Var(origine)}$$ is the coefficient of the regression of education on origin.

As $\beta_{educ} > 0$ and $\delta < 0$ for disadvantaged groups, the bias is negative : we over-estimate in absolute value the wage's disadvantage of these groups when we omit education. In this case, we attribute the disadvantage for the origins, where it is partly because these individuals have less education levels, in mean. When we control by education, disadvantage due to origin decreases. 

### Question 2b

We can choose the variables csp_pere and csp_mere (socio-professional category of the father and mother) as instrumental variables for education. Rather than using the raw CSP codes directly, we regroup them into five categories following the standard French PCS nomenclature, plus a category for non-stated CSP. 
We construct csp_pere_cat and csp_mere_cat as follows : 
- category 1 (farmers : codes 10, 11, 12, 13, 71)
- category 2 (craftsmen and business owners : codes 21, 22, 23, 72)
- category 3 (senior managers and professionals : codes 31, 33, 34, 35, 37, 38, 74)
- category 4 (intermediate professions : codes 42, 43, 44, 45, 46, 47, 48, 75)
- category 5 (employees and manual workers : codes 52–56, 62–69, 77, 78), used as the reference category as it is the most frequent
- category 0 (not stated or without professional activity : codes 00, 81, 82).

We assign retired individuals to their former occupational category.
Both variables are treated as unordered factors with category 5 as the reference, generating five dummies per parent, for a total of 10 instruments.

```{r}
lastq_emp <- df %>%
  filter(t == 6, acteu == "Employed", !is.na(salhoraire))

cat("N at the beginning :", nrow(lastq_emp), "\n")

lastq_emp <- lastq_emp %>%
  mutate(
    # CSP of father as a categorical variable
    csp_pere_cat = case_when(
      csp_pere %in% c("31","33","34","35","37","38","74") ~ 3,
      csp_pere %in% c("21","22","23","72")                ~ 2,
      csp_pere %in% c("42","43","44","45","46","47","48","75") ~ 4,
      csp_pere %in% c("52","53","54","55","56","77")      ~ 5,
      csp_pere %in% c("10","11","12","13","71")           ~ 1,
      csp_pere %in% c("62","63","64","65","67","68","69","78") ~ 5,
      TRUE ~ 0
    ),
    csp_mere_cat = case_when(
      csp_mere %in% c("31","33","34","35","37","38","74") ~ 3,
      csp_mere %in% c("21","22","23","72")                ~ 2,
      csp_mere %in% c("42","43","44","45","46","47","48","75") ~ 4,
      csp_mere %in% c("52","53","54","55","56","77")      ~ 5,
      csp_mere %in% c("10","11","12","13","71")           ~ 1,
      csp_mere %in% c("62","63","64","65","67","68","69","78") ~ 5,
      TRUE ~ 0
    )
  )

# Conversion using a reference factor of 5 (office staff/manual workers)
lastq_emp$csp_pere_cat <- relevel(as.factor(lastq_emp$csp_pere_cat), ref = "5")
lastq_emp$csp_mere_cat <- relevel(as.factor(lastq_emp$csp_mere_cat), ref = "5")

# Verifications
freq(lastq_emp$csp_pere_cat)
freq(lastq_emp$csp_mere_cat)

cat("final N :", nrow(lastq_emp), "\n")
```

Discussion of the validity of the instruments;
We must respect 2 conditions for an instrumental variable : 

1) Relevance condition (i.e. cov(instrument, education) not equal to 0)
In the sociological and economic literature, we know that parental socio-professional category is highly correlated with the educational attainment of children. Children of senior managers and professionals have substantially higher probabilities of attaining higher education, while children of manual workers have substantially lower probabilities. 
Empirically, we will confirm that by the first-stage F-statistic

2) Exogeneity condition (i.e. cov (instrument, residuals) = 0)
This condition requires that parental category socio-professionnal affects the individual's wage only through the education channel, and not directly.

The respect of this condition is more debatable, because it remains social capital transmission : for example, children of senior managers may benefit from parental networks, job referrals, or behavioural norms that directly improve their labour market outcomes independently of the diploma they hold. In that case, the instrument is invalid.

To identify a model with instrumental variables, the number of instruments must be superior or equal to the number of endogeneous variables. Here, we have only one variable endogeneous (education), and 10 instruments. So, the model is overidentified, that improves first-stage precision.

### Question 2c

We treat education as a continuous variable (education_num, coded from 1 for the highest level to 6 for no diploma), which allows us to instrument a single endogenous variable rather than five dummies simultaneously.

```{r}

lastq_emp <- lastq_emp %>%
  mutate(
    education_num = case_when(
      education == ">Bac+2"  ~ 1,
      education == "Bac+2"   ~ 2,
      education == "Bac"     ~ 3,
      education == "CAP-BEP" ~ 4,
      education == "Brevet"  ~ 5,
      education == "Aucun"   ~ 6,
      TRUE ~ NA_real_
    )
  )

# Filter on education_num 
lastq_emp <- lastq_emp %>%
  filter(!is.na(education_num))

```
We make the first stage step by regressing education_num on instrumental variables and other control variables :  
```{r}
first_stage <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif +
    csp_pere_cat + csp_mere_cat,
  se_type = "stata",
  data = lastq_emp
)
modelsummary(first_stage, stars = TRUE, fmt = 3)

```
As the model is  overidentified with 9 overidentifying restrictions, we perform a Sargan-Hansen test of instrument validity. The joint relevance of the instruments is assessed via a Wald test comparing the full first-stage model to a restricted model excluding all parental CSP dummies.
```{r}
first_stage_restricted <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif,
  se_type = "stata",
  data = lastq_emp
)
waldtest(first_stage, first_stage_restricted)

# F-stat : we need to compute it because it is not done with the function lm_robust
r2_full <- summary(first_stage)$r.squared
r2_res  <- summary(first_stage_restricted)$r.squared
n <- nrow(lastq_emp)
k <- length(coef(first_stage))
q <- 10  # number of dummies instruments (4 father + 4 mother, cat0 to cat4)
F_stat <- ((r2_full - r2_res) / q) / ((1 - r2_full) / (n - k))
cat("F-stat instruments :", round(F_stat, 2), "\n")
```
The relevance condition requires that the parental CSP level variables are correlated with the individual's education level, conditional on the other covariates ; that we can assess through the first-stage regression.

We can see that, relative to the reference category (employees and manual workers, category 5), children of senior managers and professionals have higher educational attainment, as reflected by the large negative coefficients on csp_pere_cat3 (−0.634, p < 0.001) and csp_mere_cat3 (−0.437, p < 0.001).
Children of intermediate profession parents also attain higher education levels (csp_pere_cat4 : −0.340, p < 0.001 ; csp_mere_cat4 : −0.374, p < 0.001).
By contrast, the coefficients on agricultural workers (category 1) and self-employed mothers (category 2) are not individually significant, suggesting that children from these backgrounds have educational attainment similar to the reference group.

The test yields Chi²(10) = 286 (p < 0.001), equivalent to an F-statistic of 24,06, well above the conventional threshold of 10 used to detect weak instruments.

The null hypothesis of jointly irrelevant instruments is therefore strongly rejected, and the relevance condition is satisfied.

We are now performing the 2SLS regression : 
```{r}
iv_model <- iv_robust(
  logsalhoraire ~ homme + age + age2 + origine + csp_actif + education_num |
                  homme + age + age2 + origine + csp_actif +
                  csp_pere_cat + csp_mere_cat,
  se_type = "stata",
  data = lastq_emp
)
summary(iv_model, diagnostics = TRUE)
```

### Question 2d

We will compare the OLS estimator of the coefficient of education_num in the OLS regression m6 ; and the 2SLS coefficient of the last regression : 
```{r}
# m6 0LS regression with education_num 
m6_num <- lm_robust(
  logsalhoraire ~ origine + education_num + homme + age + age2 + csp_actif,
  se_type = "stata",
  data = lastq_emp
)
summary(m6_num)

# comparison between OLS and 2SLS
library(texreg)
screenreg(
  list(m6_num, iv_model),
  custom.model.names = c("OLS", "2SLS"),
  digits = 4
)
```

The 2SLS estimates confirm what we had found with the OLS specification. Correcting for the endogeneity of education increases the estimated return to education in absolute value, from −0.052 in OLS to −0.065 in 2SLS. This is consistent with the downward bias argument (Question 2a) : OLS underestimates the causal return to education because unobserved factors simultaneously reduce both educational attainment and wages. After correction, each additional step down in the education ladder is associated with a 6.5% reduction in hourly wages.

Regarding origin, the results are stable across both specifications : only Southern European origin displays a statistically significant positive wage premium (+5.7%, p < 0.05), while no other origin group shows a significant wage gap relative to French-origin individuals. This confirms that correcting for the endogeneity of education does not substantially alter our conclusions about wage inequalities by origin. Observable factors such as educational attainment and occupational category account for the main part of the gross wage gaps identified in question 1; and the data show no evidence of direct wage discrimination for the same job and qualifications.

However, the exogeneity condition for the instruments (parents’ socio-economic status) remains impossible to verify directly. If parents’ socio-economic status transmits social capital beyond education, the 2SLS estimates remain biased. 

## Question 3
### Question 3a

We estimate a pooled OLS model on the panel sub-sample of employed individuals observed at t=1 and t=6, with clustered standard errors at the individual level to account for serial correlation within individuals. We make the random effects assumption that individual effects ($\alpha_i$) are uncorrelated with the regressors.

```{r}
# We create the panel subset for active individuals for t=1 and t=6 

panel_q3 <- df %>%
  filter(t %in% c(1, 6),
         acteu == "Employed",
         !is.na(salhoraire)) %>%
  mutate(
    # We create temporal dummies : 1 if t=6, 0 if t=1

    d_last = as.numeric(t == 6),
      )

cat("N observations :", nrow(panel_q3), "\n")
cat("N individuals :", n_distinct(panel_q3$id_individu), "\n")


# Pooled OLS model
# We create a temporal dummy d_last to catch the common effect of the period
pooled_ols <- lm(logsalhoraire ~ origine + education + homme + age + age2 + csp_actif + d_last, data = panel_q3)

# Clustered standard errors at the "id" level
clustered_se <- vcovCL(pooled_ols, cluster = ~id_individu)

# Summary of the model with clustered standard errors
coeftest(pooled_ols, vcov = clustered_se)
```
d_last coefficient is +0.015 (p = 0.003). This means that, between the first and last waves observed, wages increased by an average of 1.5% all other things being equal. This captures the conjunctural effect common to all individuals.

Education remains the strongest determinant of wages, with a significant gradient across levels. 
Among origin groups, only Southern European origin shows a significant positive coefficient (+4.5%, p < 0.05)
 
Finally, the results are consistent with the cross-sectional estimates of Question 1.

### Question 3b

We can include the origine variable in the pooled OLS model estimated in the previous question because this estimator does not eliminate time-invariant regressors (such as the origin).

However, the pooled OLS estimates do not resolve potential endogeneity problems because it relies on an unrealistic assumption. Indeed, the consistency of OLS requires that the individual effects ($\alpha_i$) are uncorrelated with all regressors, including origine. However, this assumption is unlikely to hold. Indeed, the individual effect ($\alpha_i$) captures all time-invariant unobserved characteristics of individuals, such as unobserved ability or family cultural capital, that simultaneously influence both the individual's labour market position and their origin group. We can suppose that individuals with an immigrant backgrounds may differ from French-origin individuals in unobserved dimensions (social networks and cultural norms), that directly affect wages. These factors would be absorbed into ($\alpha_i$), creating a correlation between ($\alpha_i$) and origine. 

As a result, the pooled OLS suffers from an omitted variable bias and does not consistently estimate the causal effect of origin on wages.


### Question 3c

When we allow individual effects ($\alpha_i$) to be correlated with the regressors, we can use methods such as the within estimator and the first-difference estimator. Nonetheless, these methods eliminate $\alpha_i$ because they exploit within-individual variation over time. Then, any time-invariant variable is also eliminated in the process. 

However, the variable origin is unfortunately time-invariant by construction, because parental nationality at birth does not change over time.

So, with the within estimator and the first-difference estimator, coefficients of origin are not identified. Therefore, we can not simultaneously correct for the correlation between individual effects and regressors, and identify the causal effect of origin on wages.

# 2. Extensive Margin (Employed vs. Unemployed)

## Question 4
### Question 4a

In this model, we assume that the unobserved individual heterogeneity ($\alpha_i$) is uncorrelated with the explanatory variables ($E(\alpha_i | X_{it}) = 0$). To estimate this, we need to build a panel data frame containing all active individuals across all available periods.


The primary advantage of the linear probability model is that its estimated coefficients directly represent the average marginal effect of the regressors on the probability of unemployment. Non-linear models face significant structural limitations in panel data settings, most notably the incidental parameters problem and severe complications when integrating instrumental variables (IV) for dynamic models. Although the linear probability model produces heteroskedasticity and can yield predicted probabilities outside the logical [0, 1] bounds, the issue is addressed by computing clustered robust standard errors.

```{r}
# We create the panel subset for active individuals across all periods
#If an individual is active at a period but inactive in another, we simply don't count the periods where the individual is inactive
panel_final <- df %>%
  mutate(
    unemp = case_when(
      acteu == "Unemployed" ~ 1,
      acteu == "Employed"   ~ 0,
      acteu == "Inactive"   ~ NA_real_ 
    ),
    csp_cat = ifelse(csp_actif == "" | csp_actif == " ", NA, as.character(csp_actif)),
    csp_cat = as.factor(csp_cat),
        origine_cat = as.factor(origine),
    homme_cat = as.factor(homme),
    age = as.numeric(age),
    age2 = age^2
  )

p_active <- pdata.frame(panel_final, index = c("id_individu", "t"))

lpm_re <- plm(unemp ~ origine_cat + homme_cat + age + age2 + csp_cat, 
               data = p_active, 
               model = "random")

summary(lpm_re)

# We compute clustered standard errors at the individual level to account for serial correlation
coeftest(lpm_re, vcov = vcovHC(lpm_re, type = "HC1", cluster = "group"))
```
About the specification of our model :

- We assigned a missing value (NA) for the unemployment indicator for individuals who are inactive, rather than being coded as employed or unemployed. We consider that inactivity is a distinct labour market status, and including inactive individuals as either employed or unemployed would misrepresent their situation. As a result, only periods during which an individual is active (employed or unemployed) contribute to the estimation

- The socio-professional category "Unemployed and never worked" is retained in the model but its coefficients don't have economic interpretation : by construction, an individual who has never worked cannot be employed, making the correlation between this category and unemployment status largely mechanical and uninformative from an economic standpoint.

- The selection of explanatory variables is guided by the "all else equal" principle, aiming to isolate the specific impact of origin by controlling for standard labor market determinants. The variable origine_cat is the primary variable of interest, constructed from parents' nationalities at birth to capture potential inequalities or discrimination linked to ancestry.

- To ensure the validity of this measure, several control variables are included. Gender (homme_cat) is a fundamental demographic control, as labor market participation and unemployment risks often differ structurally between men and women. The inclusion of age and its quadratic term age2 accounts for the non-linear effect of professional experience; unemployment probability typically decreases as individuals gain experience but may rise again for older workers nearing the end of their careers.

- The socio-professional category (csp_cat) is included to control for the individual's position within the labor market hierarchy. Since different sectors and skill levels face varying degrees of job stability, controlling for CSP ensures that the estimated effect of origin does not merely reflect a concentration of certain groups in more precarious occupations.

The analysis employs a Linear Probability Model with Random Effects to estimate the determinants of unemployment. Because the dependent variable is binary, the estimated coefficients can be directly interpreted as the change in the probability of being unemployed. The Random Effects specification relies on the assumption that unobserved individual heterogeneity—such as innate characteristics that remain constant over time—is strictly uncorrelated with the explanatory variables in the model. 

Since the Linear Probability Model generates heteroskedasticity and the panel structure implies serial correlation among observations of the same individual, we use robust standard errors clustered at the individual level to ensure valid statistical inference. Regarding the primary variable of interest, the origin of the individuals, the reference category is set to individuals of French origin. 

The results reveal a significant labor market penalty for several groups. Holding age, gender, and socio-professional category constant, individuals of Northern African origin face a probability of being unemployed that is 10.87 percentage points higher than their French-origin counterparts, an effect that is highly statistically significant. Similarly, individuals originating from the Rest of Africa and Southern Europe exhibit higher probabilities of being unemployed, by 8.01 and 3.75 percentage points respectively, both significant at the 5% level. The coefficients for other origins, namely Eastern Europe, Northern Europe, the Middle East, and the Rest of the World, do not show statistically significant differences from the reference group.

Gender plays a significant role, with men exhibiting a 2.28 percentage point lower probability of being unemployed compared to women. The effect of age is captured through both linear and quadratic terms. 

The effect of age follows a U-shaped profile. The negative coefficient on age (−0.0075, p < 0.001) indicates that the probability of unemployment decreases as individuals accumulate experience, reflecting the value of professional networks. The positive coefficient on age2 (+0.000071, p < 0.001) captures the reversal of this trend for older workers, whose unemployment risk rises again as they approach retirement age and face potential discrimination or skill obsolescence.


Socio-professional categories strongly correlate with employment status. The coefficient on csp_catUnemployed never worked (+0.835, p < 0.001) reflects a mechanical relationship rather than a causal one : by construction, individuals in this category have never held a job and are therefore recorded as unemployed by definition. This coefficient should not be interpreted as an economic effect.
Among the economically meaningful categories, manual workers face a significantly higher unemployment probability (+4.08 percentage points, p < 0.05) relative to the reference group (CPIS, senior managers), while craftsmen (−4.60 percentage points) and farmers (−4.48 percentage points) face lower unemployment risks, though these effects are only marginally significant once standard errors are clustered.

Despite these findings, the core assumption of the Random Effects model (that unobserved individual effects are uncorrelated with the regressors) remains a strong and potentially restrictive premise. It is highly plausible that unobserved traits, such as personal motivation or social networks, are correlated with an individual's origin or socio-professional category, thereby introducing an omitted variable bias into the current estimates. While a Fixed Effects (Within) model would traditionally correct for this endogeneity by eliminating time-invariant unobserved heterogeneity, implementing it in this context poses a structural issue: the Within transformation would simultaneously eliminate all time-invariant explanatory variables, thereby making it impossible to identify the causal effect of origin.


### Question 4b


To address the potential endogeneity problems and allow for correlation between individual effects and the regressors, a Fixed Effects or Within model is considered. This framework is theoretically superior for resolving omitted variable bias because it rigorously controls for all unobserved, time-invariant individual heterogeneity. However, applying this method to study the effect of the origin variable reveals a fundamental structural limitation. The Within transformation operates by subtracting the individual-specific time average from each observation to eliminate the unobserved fixed effect. Because demographic characteristics such as origin and gender are strictly time-invariant for a given individual across the six observed quarters, their value at any given time $t$ is exactly equal to their individual time average. The within-group differencing process perfectly cancels these variables out, resulting in a value of zero for all observations and rendering the estimation of their coefficients impossible due to perfect collinearity with the fixed effects.

The practical implementation of this model in the current dataset is hampered by the availability of data during transitions between labor market statuses. Variables related to the labor market, such as the socio-professional category, are often only defined or available when an individual is active, leading to missing values during periods of inactivity. Therefore, while the Fixed Effects model is more robust to endogeneity stemming from unobserved traits, it completely precludes the identification and estimation of the causal effect of origin on the probability of being unemployed. This structural constraint ultimately justifies the continued reliance on the Random Effects framework to quantify the labor market penalty associated with an individual's origin, despite the more restrictive assumption of exogeneity regarding the individual effects.

## Question 5
### Question 5a

Strict exogeneity requires $E(Unemp_{it'}*\epsilon_{it})=0$ for all t,t′. In a dynamic model where $X_{it}=Unemp_{it−1}$, this assumption is not credible. Indeed, $Unemp_{it-1}$
depends on $\epsilon_{it-1}$, which is correlated with $\epsilon_{it}$ if shocks are serially correlated. Moreover, future values of unemployment $Unemp_{it+1}$ depend on current shocks $\epsilon_{it}$, further violating strict exogeneity.

Weak exogeneity only requires $E(Unemp_{is}*\epsilon_{it})=0$ for s<t, which means that past unemployment is uncorrelated with current shocks. This is more credible if the error terms $\epsilon_{it}$ are serially uncorrelated. Under weak exogeneity, past levels $Unemp_{it-2}$ are valid instruments for $\Delta Unemp_{it-1}$ in the first-difference equation, which is the strategy we implement : 

```{r}
# We create the lagged variables and differences taking the panel structure into account
p_active_dyn <- p_active %>%
  group_by(id_individu) %>%
  arrange(t) %>%
  mutate(
    # Lagged levels
    l.unemp = dplyr::lag(unemp, 1),
    l2.unemp = dplyr::lag(unemp, 2),
    
    # First differences
    d.unemp = unemp - l.unemp,
    d.l.unemp = l.unemp - l2.unemp # This corresponds to \Delta Y_{t-1}
  ) %>%
  ungroup()

# We estimate the First-Differenced equation using IV 
# We instrument the differenced lag (d.l.unemp) with the twice-lagged level (l2.unemp)
dyn_iv <- ivreg(d.unemp ~ d.l.unemp | l2.unemp, data = p_active_dyn)

summary(dyn_iv)
coeftest(dyn_iv, vcov = vcovCL(dyn_iv, cluster = ~id_individu))

```
In this question, we need to have individuals as active in all periods. To calculate the lag t - 1 and t - 2, an individual needs to be active in at least 3 consecutive periods of time for the model to run, which is not possible if we use certain trimesters only.

The estimation of a dynamic panel model is required to address the presence of state dependence, which is the causal impact of past unemployment on current labor market status. While the Fixed Effects (Within) model discussed in previous sections can control for unobserved individual heterogeneity, it fails in a dynamic context due to the "Nickell bias." This bias arises because the lagged dependent variable is mechanically correlated with the individual-specific mean of the error term, rendering the Within estimator inconsistent, especially in panels with a small number of time periods. 

Similarly, a simple First-Difference model remains biased because the differenced lagged variable ($\Delta Unemp_{it-1} = Unemp_{it-1} - Unemp_{it-2}$) is inherently correlated with the differenced error term ($\Delta \epsilon_{it} = \epsilon_{it} - \epsilon_{it-1}$), as both share a common component from period $t-1$.

To overcome these inconsistencies, we utilize Instrumental Variables (IV) within a First-Difference framework. This model is pertinent because it allows us to purge the time-invariant individual effects ($\alpha_i$) while using the twice-lagged level ($Unemp_{it-2}$) as a valid instrument for the differenced lag. This instrument is ideal as it is correlated with the past change in unemployment but uncorrelated with the current innovation in the error term, thus providing a consistent estimate of the "scarring effect."

Regarding the exclusion of variables like gender, origin, and age from the instrumentation process, their omission is a direct result of the First-Difference transformation. Variables such as origin and gender are strictly constant over time for each individual; therefore, their first difference is zero, causing them to drop out of the model entirely. While age is technically time-varying, it increases by exactly one unit in every period for every individual in the panel. This makes the first difference of age a constant (equal to 1), which is perfectly collinear with the model's intercept and is consequently absorbed by it. 


### Questions 5b

The estimation of this dynamic model yields a coefficient of 0.395 for the lagged differenced unemployment variable ($\Delta Unemp_{it-1}$), which is highly statistically significant at the 0.1% level. This positive and significant estimate provides strong empirical evidence of true state dependence in the labor market. It implies that the experience of being unemployed in the previous period significantly increases the probability of remaining unemployed in the current period, even after purging the model of time-invariant individual fixed effects such as origin, gender, or innate ability through first-differencing.

In terms of economic magnitude, a coefficient of 0.394 suggests that roughly 39.4% of the previous period's change in unemployment status persists into the current period. This result highlights a "scarring effect," where the mere fact of being unemployed negatively impacts future employment prospects, potentially due to the depreciation of human capital, loss of professional networks, or negative signaling toward employers.It is also important to note the robustness of this finding, as the Wald test and the clustered t-test confirm that the relationship is not due to random noise. 

We note that when clustered standard errors are applied at the individual level, the t-statistic drops from 18.9 to 9.2 ; so the standard errors are approximately twice as large once clustering is accounted for. This shows the importance of correcting for serial correlation within individuals. Despite this correction, the coefficient remains highly significant.

Finally, the negative R-squared (−0.415) is an artefact of the 2SLS estimator and does not indicate a model misspecification. Indeed, unlike OLS, the 2SLS estimator is not constrained to improve upon the unconditional mean prediction, so negative R-squared values are expected.



## Question 6

### Question 6a

In order to consider a model with a fixed effect, it is necessary to adapt it to the case of a binary variable. Suppose there exists a continuous variable $Y^*$, the latent variable, such that $Y=\mathbf{1}\{Y^* \geq 0\}$ and suppose that $Y^*$ is generated by a linear model with a fixed effect : 

$$
Y_{it} = X_{it}'\beta_0 + \alpha_i + \varepsilon_{it}
$$
where $\varepsilon$ is independent of $X$ and has distribution function $F$. Then
$$
E(Y_{it} \mid X_{it}, \alpha_i) \\
= P(Y_{it}=1 \mid X_{it}, \alpha_i) \\
= P(X_{it}\beta + \alpha_i + \varepsilon_{it} \geq 0 \mid X_{it}, \alpha_i) \\
= P(-\varepsilon_{it} \leq X_{it}\beta + \alpha_i \mid X_{it}, \alpha_i) \\
 \\
= F(X_{it}' \beta + \alpha_i)
$$

If we try to apply the first differences method, we obtain : 
$$
E(Y_{it} - Y_{i,t-1} \mid X_{it}, X_{i,t-1}, \alpha_i)  \\
= E(Y_{it} \mid X_{it}, X_{i,t-1}, \alpha_i) - E(Y_{i,t-1} \mid X_{it}, X_{i,t-1}, \alpha_i) \\
= F(X_{it}' \beta + \alpha_i) - F(X_{i,t-1}' \beta + \alpha_i)
$$

Yet, the function F is not linear, as it is a distribution function of a random normal variable or the logistic distribution function. This does not allow us to eliminate the fixed effect as the linear model, in order to estimate $\beta$ without biais. In conclusion, we cannot use the first differences panel model. 

If we try to apply the within panel method, we obtain : 
$$
E\big(Y_{it} - \bar{Y}_i \mid X_{it}, \alpha_i \big)\\ 
= E\big(Y_{it} \mid X_{it}, \alpha_i \big) - E\big(\bar{Y}_i \mid X_{it}, \alpha_i \big) \\
= E\big(Y_{it} \mid X_{it}, \alpha_i \big) - E(\frac{1}{T} \sum_{t=1}^{T}{Y}_{it} \mid X_{it}, \alpha_i \big) \\
= E\big(Y_{it} \mid X_{it}, \alpha_i \big) - \frac{1}{T}\sum_{t=1}^{T}E({Y}_{it} \mid X_{it}, \alpha_i \big) \\
= F(X_{it}' \beta + \alpha_i) - \frac{1}{T} \sum_{t=1}^{T} F(X_{it}' \beta + \alpha_i)
$$
Yet, the function F is not linear, as it is a distribution function of a random normal variable or the logistic distribution function. This do not allow us to eliminate the fixed effect as the linear model, in order to estimate $\beta$ without biais. In conclusion, we cannot use the within panel model. 

### Question 6b 

First, we create the correct subset to restrict our analysis to cross-sectional data using the last quarter and focusing on active individuals (employed or unemployed). We create then the indicator of being unemployed. 
```{r}
# subset 
lastq <- df %>%
  dplyr::filter(t == 6, acteu != "Inactive") 
cat("N =", nrow(lastq), "active individuals")

# recoding for modeling 
lastq <- lastq %>% mutate(unemp = ifelse(lastq$acteu == "Unemployed", 1, 0))

# Check
freq(lastq$unemp)
freq(lastq$acteu)
```
Our analysis will hold on 5030, including 324 individuals who are unemployed. We suppose that the variable csp_actif in the data followed the french INSEE convention : for the unemployed individuals, the csp_actif indicates the CSP associated with their last employment. 

First, we have to clean the data.  
```{r}
#freq(lastq$csp_actif) # modality empty ("") without any individualis in it 
lastq$csp_actif_clean <- droplevels(lastq$csp_actif)
levels(lastq$csp_actif_clean) # check 

# last check 
table(lastq$csp_actif,lastq$csp_actif_clean) # only terms in the diagonal: OK 
```

#### Probit model 

We begin by estimating a probit model. 
```{r}
probit_model <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + I(age^2), family = binomial(link = "probit"), data = lastq)
summary(probit_model) # we interpret the sign and the significance of the coefficients, but not their values 
```
The model's AIC is 2208,2. And we can only interpret the coefficient qualitatively and its significance. So, we will concentrate on the interpretation of the marginal effect. 
- The Northern African origin has a positive effect on the probability of being unemployed all else being equal, and is statistically significant at 1%.   
- The Eastern European origin has a positive effect on the probability of being unemployed all else being equal, and is statistically significant at 10%. 

We compute the estimated Average Marginal Effect (AME) in the probit model. 
```{r}
AME_probit <- margins(probit_model)
summary(AME_probit)
```
Note that the variable origin is a categorical variable. Formally, when we add this variable in this type of model, we include all the indicators for the categories, except one that is the category of reference (France in this case), as in a linear model. Analyzing the average marginal effect of the variable origin means to look at the AME of each dummy variable, in reference to the French origin. 

Indeed, if we denote \(X_j\), one of the dummies variable origin, the average effect of switching \(X_j\) from 0 to 1 is then:

$$
E\Big[ F(X_{-j}' \beta_{0,-j} + \beta_{0j}) - F(X_{-j}' \beta_{0,-j}) \Big]
$$
In addition, in a binary model, the units of the AME are points of percentage. We will denote this unit "pp". 

As it was sid previoulsy, only one origin has a significant effect at 1% on the probability of being unemployed : it is the Northern Africa (Maghreb) origin. All else being equal, to have at least one parent Northern African descent rather than having both French parents increases of 9,2 pp on average the probability of being unemployed. But, it is important to underline that the AME of the Eastern europe origin is not significant at 10%, whereas it coefficient is, meaning that the magnitude of the change in probability caused by this variable statistically different from zero, even if it seems plausible that the variable have a significant qualitative effect. This difference in statistical significance means we must be cautious about our conclusions, and it is best not to state with certainty that this variable has a significant effect.  

Note the other variables of origin are not statistically significant. So, there is no evidence of inequality in employment, taking into account these covariate.  

#### Logit model

We now estimate a logit model. 
```{r}
logit_model <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + I(age^2), family = binomial(link = "logit"), data = lastq)
summary(logit_model) # we interpret the sign and the significance of the coefficients, but not their values 
```
The model's AIC is 2372,3, instead of 2208,2 for model 1. According to this measure, the probit model is better than the logit model. 

We compute the Average Marginal Effect (AME) in the logit model. 
```{r}
AME_logit <- margins(logit_model)
summary(AME_logit)
```
The result are quite identical with the ones of the probit model. Only the Northern African origin is statistically significant, and it is at 1%, and the estimated AME is a bit more important than in the probit model : 9,3 pp rather than 9,2 pp. 

#### Different specifications

To be able to observe some differences comparing two non-French origins, we test some models behind.  
```{r}
lastq$origine_cat <- relevel(lastq$origine_cat, ref = "Middle East")
logit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "logit"), data = lastq)
summary(margins(logit_model2))
```
No variable has a significant effect on the probability studied. 

```{r}
lastq$origine_cat <- relevel(lastq$origine_cat, ref = "Rest of Africa")
logit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "logit"), data = lastq)
summary(margins(logit_model2))
```
No variable has a significant effect on the probability studied. 

```{r}
lastq$origine_cat <- relevel(lastq$origine_cat, ref = "Maghreb")
logit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "logit"), data = lastq)
summary(margins(logit_model2))
```
To see some differences, we try to compare the effect of being from another origin than Northern African ones, has it was the only one to be significant previously. Except French people, the individuals which have less chance of being unemployed in average, all else being equal, than the individuals with Northern African origins are : 
- the one with Northern European origins (-9,1 pp, significant at 5%)
- the ones with Southern European origins (-7,8 pp, significant at 1%)
- the ones with "Rest of the world origin" ("Rest of the world origin" + Southern Asia) (-11,3 pp, significant at 1%). 
The effect of the other origins cannot be stated as statistically different than the Northern African ones. 

This provides a more in-depth analysis of the effect of origins in the labor market and the access to employment.  

### Question 6c 

We create the indicator of having a higher education degree (modality "1" of the variable education in the dataset). 
```{r}
lastq$origine_cat <- relevel(lastq$origine_cat, ref = "France")
lastq <- lastq %>% mutate(highdegree = ifelse(lastq$education == ">Bac+2", 1, 0))
# check
#freq(lastq$highdegree) # 975 individuals
#freq(lastq$education)
```

Before using the Rivers-Vuong method, we include the new indicator as an explanatory variable, together with "origin" and the other relevant covariates, using a probit model. We choose the probit model because of the differences of model's AIC. 
```{r}
probit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + I(age^2) + highdegree, family = binomial(link = "probit"), data = lastq)
summary(probit_model2) # we interpret the sign and the significance of the coefficients, but not their values 
AME_probit2 <- margins(probit_model2)
summary(AME_probit2) ; summary(AME_probit) # quantitative interpretation is possible 
```

We analyse quickly the result of the regression, with the indicator of having a higher education degree in the covariates. The age is significant at the 10% level, whereas the sex is not. The covariates of the CSP are significant are 1% (except for the not stated). The new indicator is not significant statistically The AME of the indicator of being from Norther African origin is a bit more important in this model than in the previous one : all else being equal, to have at least one parent with a nationality from Northern Africa rather than having both French parents increases of 9,3 pp on average the probability of being unemployed. The difference is only of 0,1 pp, which may be not significant. 


For the same reasons than in question 2, controlling for education level may raise an endogeneity problem. 

An individual's level of education, particularly obtaining a higher education degree, results in part from their own choices, which are influenced by unobservable characteristics (such as family cultural capital, family social capital, ability, motivation, …). These characteristics also directly affect the chance of being employed. For example, the family social capital and the strength of weak ties make it easier to find a job, thanks to the information circulating within one’s network of people with significant economic and cultural capital. This network, though for example role model, influence the probability of having a higher degree education. Furthermore, finding a job rather than remaining unemployed may also depend on the motivation and effort put into the job search, much like the motivation required to obtain a degree. 

Furthermore, origin and education are correlated. Individuals from certain origin groups have on average lower educational attainment, due to inequalities in access to education. Then, part of the effect of origin on the probability of being employed operates through the higher education degree channel (origin affects education, which in turn affects the probability of being employed). 
  
In the following treatment, we will ignore the fact that the Rivers-Vuong method does not work for a binary endogeneous variable and we will relax the assumptions regarding the laws of certain variables. We apply this method, with the same instrument than before, the CSP of the father. As we only have a binary variable to instrument, we chose to restrict the number of instruments comparing to part 1.  The instrument have to fulfill 2 conditions : 

1) Relevance condition (i.e. cov(instrument, highdidegree) not equal to 0)
In the sociological and economic literature, we know that parental socio-professional category is highly correlated with the educational attainment of children, and more precisely on the chance of having a higher education degree. Children of senior managers and professionals have substantially higher probabilities of attaining higher education, while children of manual workers have substantially lower probabilities. 

2) Exogeneity condition (i.e. cov (instrument, residuals) = 0)
This condition requires that parental category socio-professional affects the individual's wage only through the education channel, and not directly.


As the variable highdegree is potentially endogeneous, we consider this model : 
$Y=\mathbf{1}\{X'_1\beta_1 + X_2\beta_2+ \epsilon \geq 0\}$, where is $X_1$ the matrix of the control variables (homme, age, age2) and $$X_2$$ is the indicator $highdegree$, which is endogeneous, meaning here not independent from $\epsilon$. 

We choose the same instrument $Z$ that before : the csp of the father. It supposed to have a direct effect on highdiplom but not on $Y$, meaning that we supposed that : 

$$ 
X_2 = X'_1\gamma_1 + Z\gamma_2 + \eta
$$

where $\epsilon = \rho\eta + \zeta $

The model is now : 
$$
Y=\mathbf{1}\{X'_1\beta_1 + X_2\beta_2+ \rho\eta + \zeta \geq 0\}
$$

We create the variable of the regrouped variable of the csp_pere as in the question 2. The instrument is categorical. 
```{r}
lastq <- lastq %>% mutate(csp_pere_cat = case_when(
    csp_pere %in% c("31","33","34","35", "37","38", "74")          ~ 3,  # Managers and former managers
    csp_pere %in% c("21","22","23", "72")                    ~ 2,  # Craftspeople and Business Owners
    csp_pere %in% c("42","43","44","45","46","47","48", "75")        ~ 4,  # Intermediate occupations
    csp_pere %in% c("52","53","54","55","56", "77")              ~ 5,  # employees
    csp_pere %in% c("10","11","12","13", "71")                 ~ 1,  # Farmers
    csp_pere %in% c("62","63","64","65","67","68","69", "78")        ~ 5,  # Blue-collar workers and former blue-collar workers
    csp_pere %in% c("00", "81", "82") ~ 0, #not stated, we regrouped the 81 and 82 here, because it is only on individuals who cannot be alone in a specific category
    TRUE                                      ~ 0)) # we regrouped the NA and the O if necessary

lastq$csp_pere_cat <- as.factor(lastq$csp_pere_cat)
freq(lastq$csp_pere_cat)
lastq$csp_pere_cat <- relevel(lastq$csp_pere_cat, ref = "5") # the most frequent one 
```


Firstly, we compute the related OLS regression to obtain en estimation of $\eta$. This corresponds to the first stage. 
```{r}
mod <- lm(highdegree ~ origine_cat + homme + csp_actif_clean + age + I(age^2) + csp_pere_cat, data = lastq, na.action = na.exclude) # we use na.exclude beacause it retains the position where data was missing in the final residual vector, which is very important here  
lastq$etahat<- mod$residuals
summary(mod)
```
Note that we use 'lm' and not 'lm_robust' because 'na.exclude' do not exist in 'lm_robust'.

We check the relevance conditions : the condition inside the indicator specifies the condition on csp_pere for the flag to be set to 1, with 0 always being the default value (5). 
- $\mathbb{1}_{\{\text{csp_pere_cat} = 0\}}$ is not significant at 10% 
- $\mathbb{1}_{\{\text{csp_pere_cat} = 1\}}$ is significant at 10% 
- $\mathbb{1}_{\{\text{csp_pere_cat} = 2\}}$ is not significant at 1% 
- $\mathbb{1}_{\{\text{csp_pere_cat} = 3\}}$ is not significant at 1% 
- $\mathbb{1}_{\{\text{csp_pere_cat} = 4\}}$ is not significant at 1% 
Only one indicator fails to meet the relevance condition: the 'not stated' category. The joint relevance of the instruments is assessed via a Wald test comparing the full first-stage model to a restricted model excluding all father CSP dummies.

```{r}
first_stage_restricted <- lm(
  highdegree ~ homme + age + I(age^2) + origine_cat + csp_actif_clean,
  data = lastq
) # we keep lm here, to use the same model than the first stage (mod). 
waldtest(mod, first_stage_restricted)
```
The relevance condition requires that the parental CSP level variables are correlated with the fact that the individual has a higher education degree or not, conditional on the other covariates.

The F-statistic of 45,2 in the Wald test is well above the conventional threshold of 10 used to detect weak instruments, knowing that this threshold is usually too small to detect weak instrument when the binary variable is highly skewed (93,6% of 0 and 6,4% of 1). The p-value is less than 1%. The null hypothesis of jointly irrelevant instruments is therefore strongly rejected, and the relevance condition is satisfied.

Then, we compute the logit model on  $Y=\mathbf{1}\{X'_1\beta_1 + X_2\beta_2+ \beta_3\widehat\eta + \zeta \geq 0\}$. Note that the variable $\zeta$ are the residuals of the regression. This is the second stage. 

```{r}
model_instru <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + I(age^2) + highdegree + etahat, family = binomial(link = "probit"), data = lastq)
AME_instru <- margins(model_instru)
summary(model_instru) ; summary(AME_instru) ; summary(AME_probit2) 
```
Note that the AME of the Northern African origin is a little bit more important in the model with the instrument (9,56 pp) than in the one without (9,29 pp). The effect can be underestimated in the model without the instrumentation. Indeed, the potential endogeneity of the variable introduced a confounding bias that underestimated this effect. The use of an instrumental variable made possible to isolate the origin effect from these factors, revealing its impact, which was slightly higher. This increase is only about 0,3 pp, meaning that this potential bias was not very strong. 

Nevertheless, we compute the Wu-Hausman test, which is the test of the significance of the coefficient of $\hat\eta$ (the test is valid in our case with the potential endogeneous variable). It is not significant ($p-value = 57\%$). As it is the coefficient of the first stage residual, we cannot reject the nullity of the coefficient, meaning that we cannot reject that our variable highdegree is not endogenous. Note that we tested the instrument with all the category of csp_pere (not regrouped), the p-value of the coefficient of the first stage residual was not significant ($p-value = 22\%$ $< 57\%$). If the instrument detailed is more precise than the regrouped one, the coefficient was still not significant. 

Here, we can do some hypothesis to understand these conclusions :
- it can be good news that the variable of the high degree is not endogeneous, but this conclusion remains unlikely. 
- or the variable of the csp_pere is not exogeneous, which is highly likely as we explained before. 

Regarding the significance of the variable origins, we can also question the number of people in the dataset : 
```{r}
table(lastq$unemp, lastq$highdegree)
table(lastq$unemp, lastq$origine_cat)
```
There are only 40 people with a higher education degree who are unemployed in this subset. In each category, there are between 5 and 31 individuals who are not employed and with an other origin than French (by conting by category of origins). Due to the specification of the probit model, these numbers are probably not large enough to have some robust conclusions about labor market discrimination, which is an analysis demanding control variable.  

We test a last time this methods on the binary variable origin, which will aggregate different effects. 
```{r}
modelbin <-  glm(unemp ~ origine_bin + homme + csp_actif_clean + age + I(age^2), family = binomial(link = "probit"), data = lastq)
AME_bin <- margins(modelbin)
summary(AME_bin)

mod2 <- lm(highdegree ~ origine_bin + homme + csp_actif_clean + age + I(age^2) + csp_pere_cat, data = lastq, na.action = na.exclude) 
lastq$etahat2<- mod2$residuals 
summary(mod) # same result than before 

model_instru2 <- glm(unemp ~ origine_bin + homme + csp_actif_clean + age + I(age^2) + highdegree + etahat2, family = binomial(link = "probit"), data = lastq)
AME_instru2 <- margins(model_instru2)
summary(model_instru2) ; summary(AME_instru2) ; summary(AME_bin) ; summary(AME_probit)
```
These results are obviously less precise, but it was useful to test this last model :
Without the instrument, all else being equal, to have at least one parent without the French nationality increases (prediction) of 3,3 pp on average the probability of being unemployed. The variable origin is statistically significant at 1%, meaning that we can potentially attest that there is discrimination on the labor market against people who are not of French descent, on average. We note that the CSP variables are not significant in this model, so we have to be careful to interpret the effect origin. Here, origin can include a bias (the "included variable" bias) : we probably capture a part of the causal effect of the CSP by including the origin as a binary variable. This can occur especially because the binary variable regroups too many different situations. The original model might be more robust than the binary one to evaluate discrimination based on the origin.  

We note that when we use the model with the instrument, we saw the same result than before. The coefficient of the first stage residual are not significant. The coefficient of the variable origin is still significant. 



