---
title: "DM Econometrics - Labor Market and Origin"
subtitle: "Econometrics 2 — ENSAE 2A — Polixène Gresant, Rachel Issiakou, Juliette Lesné"
date: "Avril 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
#install.packages()
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sandwich)    
library(lmtest)      
library(AER)         
library(plm)         
library(margins)     
library(stargazer)  
library(questionr)
library(estimatr)
library(modelsummary)
```

```{r data, include=FALSE}
url <- "https://github.com/pgresant/econo2_project/raw/main/DM_Subject_2_Data.dta"
df <- read_dta(url)

# We recode the variable to better understand them in the analysis
df$origine <- dplyr::recode(df$origine,   
  "01" = "France",
  "03" = "Nothern Europe",
  "04" = "Southern Europe",
  "05" = "Eastern Europe",
  "06" = "Maghreb",
  "07" = "Rest of Africa",
  "08" = "Middle East",
  "09" = "Southern Asia",
  "10" = "Rest of the world") 
#freq(df$origine)

df$acteu <- dplyr::recode(df$acteu,   
  "1" = "Employed",
  "2" = "Unemployed",
  "3" = "Inactive") 
#freq(df$acteu)

df$csp_actif <- dplyr::recode(df$csp_actif,   
  "1" = "Farmers",
  "2" = "Craftsmen, company head",
  "3" = "CPIS", 
  "4" = "Intermediate profession",
  "5" = "Service employees",
  "6" = "Manual workers", 
  "8" = "Unemployed never worked", 
  "0" = "Not stated"
  ) 
#freq(df$csp_actif)

df <- df %>%
  mutate(
    #contrat
    contrat = na_if(contrat, ""),  # transforms "" into NA
    contrat = dplyr::recode(contrat,
      "0" = "Not employed",
      "1" = "CDI",
      "2" = "CDD",
      "3" = "Seasonal_contract",
      "4" = "Interim",
      "5" = "Apprentissage",
      .missing = "99"), # uses the NA to recode them into 99 
      
    #education
    education = na_if(education, ""),
    education = dplyr::recode(education,   
      "1" = ">Bac+2",
      "3" = "Bac+2",
      "4" = "Bac",
      "5" = "CAP-BEP",
      "6" = "Brevet",
      "7" = "Aucun",
      .missing = "99")
    )
#freq(df$contrat)

df <- df %>% mutate(
  
  origine_bin = ifelse(origine == "France", "France", "Etranger"),
  # We make the assumption that individuals who do not state 
  # their nationalities are treated as non-French
  
  origine_bin99 = case_when(
    origine == "France" ~ "France",
    origine == "99"     ~ "Not Stated",
    TRUE                ~ "Etranger"
  )  # in the binary variable origine_bin99 we do not make the assumption that individuals whom origin is unknown are foreigners. We create a third category.
  
)

df <- df %>% mutate(origine_cat2 = case_when(
  origine == "France" ~ "France",
  origine %in% c("Rest of Africa","Maghreb", "Middle East") ~ "Africa_ME" ,
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat3 = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East",
  origine %in% c("Rest of Africa","Maghreb") ~ "Africa" ,
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat4 = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East", # 30 individuals in lastq_emp
  origine %in% c("Rest of Africa") ~ "Rest of Africa", # 61 individuals in lastq_emp
  origine %in% c("Maghreb") ~ "Maghreb", # 122 individuals in lastq_emp
  origine %in% c("Southern Europe", "Eastern Europe") ~ "Europe_S_E" ,
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))

df <- df %>% mutate(origine_cat = case_when(
  origine == "France" ~ "France",
  origine == "Middle East" ~ "Middle East", # 30 individuals in lastq_emp
  origine %in% c("Rest of Africa") ~ "Rest of Africa", # 61 individuals in lastq_emp
  origine %in% c("Maghreb") ~ "Maghreb", # 122 individuals in lastq_emp
  origine %in% c("Southern Europe") ~ "Southern Europe",
  origine %in% c("Eastern Europe") ~ "Eastern Europe",
  origine %in% c("Southern Asia", "Rest of the world") ~ "Rest of the world" ,
  origine %in% c("Nothern Europe") ~ "Nothern Europe" ,
  origine %in% c("99") ~ "Not_Stated"))


# We are going to recode some variables in the data set so that we can use them later in our project
class(df$salhoraire) # verification of the class of the dependent variables 
class(df$logsalhoraire)

vars_num <- c("age")
df[vars_num] <- lapply(df[vars_num], as.numeric) # We convert continous variables into numbers, in order to use them in regressions

vars_cat <- c("education", "origine", "contrat", "csp_actif", "csp_mere", "csp_pere", "acteu", "homme", "origine_bin99", "origine_cat", "origine_bin")
df[vars_cat] <- lapply(df[vars_cat], as.factor) # We convert categorical variables to factor, in order to use them in regressions

# We precise the category of reference
df$acteu <- relevel(df$acteu, ref = "Employed")
df$origine <- relevel(df$origine, ref = "France")
df$origine_bin <- relevel(df$origine_bin, ref = "France") # la catégorie aucun est la plus fréquente mais bon ...
df$education <- relevel(df$education, ref = "Aucun")# la catégorie 7 est la plus fréquente mais bon ...je pense on eput changer selon le subdataset
df$contrat <- relevel(df$contrat, ref = "CDI") # category la plus fréquente
df$csp_actif <- relevel(df$csp_actif, ref = "Service employees") # category la plus fréquente$
df$origine_cat <- relevel(df$origine_cat, ref = "France")
df$origine_bin99 <- relevel(df$origine_bin99, ref = "France")
df$origine_bin <- relevel(df$origine_bin, ref = "France")
# We create some variables 
df$age2 <- df$age^2
```

Some descriptive statistics in order to understand the dataset
```{r}
table(df$immi, df$nat)
```
According to INSEE, an immigrant is a person born abroad who is a foreign national and resides in France.
There is 1956 individual*quarter French people that are immigrants. Some immigrants have been able to become French citizens, whilst others remain foreign nationals. Indeed, immigrant status is permanent. 

```{r}
table(df$descimmi, df$origine)
```
People who have at least one parent that held a different nationality than the French nationality can be a descendants of immigrant(s) (their parents are immigrants, meaning that they are born abroad without the French nationality). 
The other part of people who have at least one parent that held a different nationality than the French nationality are not descendant of immigrant (their parent do not held the French nationality but are born in France). 

```{r}
cprop(table(df$immi, df$origine_bin))
```
53,2% of the individualsxquartier who have at least one parent that held a different nationality than the French nationality are immigrant, meaning they are are born abroad without the French nationality. 

We provide the correct version of the variable salhoraire. As this does not affect the results when using the logarithm as the dependent variable (except for the constant/intercept) and that we done the treatment before the correction, we will keep the original variable logsalaire.
```{r}
df$salhoraire = df$salmensuel / (df$heuretra * 4.33)
```


# 1. Intensive Margin (Net Hourly Wage)

**Question 1**

```{r q1_sample}
# We select the sample : employed individuals in the last observed quarter (t=6) 
summary(df[df$t == "6" & df$acteu == "Employed", ]$salhoraire) # 660 NA in the dependent variable

lastq_emp <- df %>%
  dplyr::filter(t == 6, acteu == "Employed",  !is.na(salhoraire)) 

cat("N =", nrow(lastq_emp), "employed individuals with observed salary \n")
summary(lastq_emp$logsalhoraire) # no NA in the dependent variable

freq(lastq_emp$annee) # 42% of the observation are made in 2015, 58% in 2016
table(lastq_emp$annee, lastq_emp$trimestre) 
prop(table(lastq_emp$annee, lastq_emp$trimestre)) 

# to take into account the effect of the time gap between the observation, we add an indicator
lastq_emp <- lastq_emp %>% mutate(indic2016 = ifelse(lastq_emp$annee == 2016, 1, 0))
freq(lastq_emp$indic2016) # OK 
```

```{r q1_stats}
# First, we make few descriptive statistics of salary 
# summary(last_q$logsalhoraire)

# Distribution of origin within employed individuals
freq(lastq_emp$origine)

# descrptive statistics 
rprop(table(lastq_emp$origine, lastq_emp$immi))
cprop(table(lastq_emp$origine, lastq_emp$immi))

```
In the subset, arounf 75% of the individuals with at least one parent with an African nationality (except Maghrebian) or a Middle East one are immigrant, whereas it's 47% for Magrhebian and Eastern Europe origins, 30% for Northern and Southern European orgins.  

This tendancy is confirmed by the second table : in the immigrant population, there 20% of people with at least one parent with an Maghrebian nationality, 17% with an african nationality and 25% with southern european nationality. 

We estimate several models, using the function lm_robust (instead of lm), to compute heteroscedasticity-robust standard errors. Indee, we are not sure that the homoscedasticity assumptions holds here.

```{r}
m1 <- lm_robust(logsalhoraire ~ indic2016 + origine, data = lastq_emp)
summary(m1) # do not put the * of significance 
modelsummary(m1, stars = TRUE) # allow us to see the * of significance  

m1.1 <- lm_robust(logsalhoraire ~ origine, data = lastq_emp)
summary(m1) ; summary(m1.1) # to compare the 2 models easily 
modelsummary(m1, stars = TRUE) 
```
The estimator of the coefficient of indic2016 is very low (<0,01) and not significant at 10%. The Ajusted R-squared is lower in the odel with the indicator than the other. So, we will not include this variable in the next models. 

In the first regression (m1.1), we regress only the logarithm of wages on origin. The coefficients that are significant at 5% of level of risk are those for the Northern European origin (+0.119), the rest of Africa (-0.177) and South Asia (-0.151). The coefficient of the origin of Middle East origin (+0,09) is significant at the 10% level, but it's less relevant than the others. This can be interpreted as follows: an employee of South Asian origin earns approximately 15% less than an employee of French origin, in average. We can underline that the coefficient of the Maghreb origin has a p-value equal to 10,12%, meaning that it is not significant at 10% level, but not so far from this level. 

Considering the effectifs and the result of the regression, we did the agregation of the variable seen upper in the code. It was necessary to agregated some categories, in order to get effectifs enough large to have be significate, when we will add other variables.
We regrouped : Southern Asia and the Rest of the world ; Easter Europe and Southern Europe, only to see the effect, and to have around 30 indivuals at least in each category. It would be possible after, to agregate the categories Southern Asia and the Rest of the world with the Not Stated (99), in terms of effect. 

```{r}
m1.2 <- lm_robust(logsalhoraire ~ origine_cat, data = lastq_emp)
summary(m1.2)
freq(lastq_emp$origine_cat)
```
These results where expected.  

```{r}
m1.3 <- lm_robust(logsalhoraire ~ origine_bin, data = lastq_emp)
summary(m1.3)
```
The dummy variable origin is significant at the 5% level. An employee of that have at least on parent who do not held the french Nationality earns approximately 3% less than an employee of French origin, in average and without variable of control. This can be interesting for us, in order to have an aggregated of this social condition, but it is not precise enough. 


```{r}
m2 <- lm_robust(logsalhoraire ~ immi + origine, data = lastq_emp)
m2.1 <- lm_robust(logsalhoraire ~ immi + origine_bin, data = lastq_emp)
m2.2 <- lm_robust(logsalhoraire ~ immi + origine_cat, data = lastq_emp)
summary(m2); summary(m2.1) ; summary(m2.2)
```
When we control by the dummy variable "immi", the origin is not . 
- In model 2.1, the variable origin is not statistically significant.  
- In model 2.2, the variable of the African origin is not significant anymore. The only significant variable are the open of Nothern European origin and Not Stated, that are more significant than before (at level 1%), and the coefficient are more important for the Northern origin (+3 point of pourcentage), meaning that the effect of this origin is broader when controlling by the immigration status. The variable of the middle East origin is significant at 10%.
- In model 2, 

Pb d'endogéniété car plusieurs corrélation entre les régresseurs. on risque ici d'ajouter un biais de variable incluse. Donc on le rajoute pa la variable immi en controle. Néanmoins, une variable d'interaction pourrait etre pertinente. 
=> ou alors on parle de cannal ntm pr les personnes africa origin??

We add an interaction variable in the model. 
```{r}
m2.3 <- lm_robust(logsalhoraire ~ immi*origine_cat, data = lastq_emp)
summary(m2.3)
```
Rien est significatif. pb effectif et biais variable incluse potentiel. on garde pas cette specification qui n'est pas pertinente. 

```{r}
m3 <- lm_robust(logsalhoraire ~ origine_cat + homme + age + age2 + contrat, data = lastq_emp)
summary(m3)
```
In this model, we regress only the logarithm of wages against gender, age, age squared, and origin. 
In absolute value, the coefficients : 
- increase for Middle East origin and Northern Europe origin variables.
- decrease for the Rest of Africa and Rest of the world origin variables.  

En termes de significativité statistique : -
- la variable d'origine du Middle East devient statistiquement significative à 10%, 
- la variable d'origine Rest of africa reste statistiquement significative à 5% (p-value of 1,06%). 
- la variable d'origine d'Europe du Nord est toujours statistiquement significative à 1%. 

```{r}
m4 <- lm_robust(logsalhoraire ~ origine + homme + age + age2 + csp_actif, data = lastq_emp)
summary(m4)
```

In this model, we regress only the logarithm of wages against gender, age, age squared, and origin. 
In absolute value, the coefficients : 
- decrease for Middle East origin, Northern Europe, and Rest of the world origin variables. origin variables.
- increase for the Rest of Africa   

En termes de significativité statistique : -
- la variable d'origine du Middle East is not statiscally significant at 10%, 
- la variable d'origine Rest of africa reste statistiquement significative à 5%. 
- la variable d'origine d'Europe du Nord reste statistiquement significative à 5%.  

The R-squared is much higjer in this model than the precedent one. 
```{r}
m5 <- lm_robust(logsalhoraire ~ homme + age + age2 + origine_cat + contrat + csp_actif, data = lastq_emp)
summary(m5)
```

Les effectifs ne sont pas assez grands pour avoir autant de variables de contrôle et conserver une inférence robuste.  

```{r}
m6 <- lm_robust(logsalhoraire ~ origine + education + homme + age + age2 + csp_actif, data = lastq_emp)
summary(m6)
```
correlation entre education et entre csp_actif (mais très agrégé)

Comparing to the model 4, in this model : 

In absolute value, the coefficients : 
- decrease for Rest of Africa, Northern Europe, Rest of the worl and Not stated origin variables. 

En termes de significativité statistique : -
- la variable d'origine du Middle East is not statiscally significant at 10%,
- la variable d'origine Rest of africa reste statistiquement significative à 10%. 
- la variable d'origine d'Europe du Nord reste statistiquement significative à 5%.  
- the variable of origin Southern Europe significant à 5%, which was not the case in the previous model. 

```{r}
m7 <- lm_robust(salhoraire ~ homme + age + age2 + origine + csp_actif, data = lastq_emp)
summary(m7)
```

**Explications**

Four different regressions are carried out to test the differences in the coefficients of the variables and their significance, depending on the variables and categories used. These are four OLS models with standard errors robust to heteroscedasticity (HC1) based on the 4 046 employees in the last quarter. In the first three regressions, the dependent variable is the log of the hourly wage; whereas in the last one, the dependent variable is the hourly wage in absolute terms. For the first three regressions, explanatory variables are added in each case to determine the channels through which origin affects wages.

In the first regression, we regress only the logarithm of wages against gender, age, age squared, and origin. The significant coefficients relating to origin are those for Northern European origin (+0.126), the rest of Africa (-0.15) and South Asia (-0.151). This can be interpreted as follows: all other things being equal, an employee of South Asian origin earns approximately 15% less than an employee of French origin.

In the second regression, the education variable is included. The origins with significant coefficients are those from Southern Europe (0.49) and the rest of Africa (-0.113). The origin coefficients decrease for almost all origins. For example, for the ‘rest of Africa’ origin, the wage gap relative to individuals of French origin is -15% in the first regression, then -11% in the second. This means that part of this wage gap is due to the fact that certain groups have, on average, a lower level of education. For those of South-East Asian origin, the wage gap with employees of French origin disappears when education is included, which therefore appears to fully explain the gross wage gap.

In the third regression, explanatory variables relating to contract type and socio-professional categories are also included. With regard to origin, the significant coefficients are those for origins associated with Northern Europe (+0.1) and Southern Europe (0.0577). The coefficients for origin diminish further and become insignificant. So another part of the pay gap can now be explained by the type of job held. The coefficient associated with African origin is no longer significant, so does the pay gap relative to individuals of French origin disappear completely once we control for the variables of education and type of job. Thus, direct discrimination regarding equal pay for equal work is not captured here, as inequality operates upstream, via access to education and good jobs. Nevertheless, there may be over-control if these variables are themselves consequences of origin.

For the final regression, we take wage as the dependent variable directly, and as a result the coefficients associated with the intercept are no longer significant. This can be explained by the high variance in wage levels, as the distribution is skewed and high wages introduce noise. In the following analysis, we will therefore use the logarithm of wage as the dependent variable;

In conclusion, wage gaps gradually narrow as we add controls. The only group for which the gap persists after controls is Southern Europe (with a coefficient of +6% in the third regression). So, at first glance, origin acts mainly through observable channels (level of education, type of employment), rather than through direct discrimination when the job and qualifications are identical.


**Question 2**
*Question 2a*
Education is a potentially endogeneous variable in the regression. Indeed, the level of education of an individual partly results from his/her own choices, which are influenced by unobservable charateristics (such as family cultural capital, ability, motivation, ...). Theses characteristics also directly affect wages. Then, we expect that the correlation between education and the residuals is not null, which violates the exogeneity assumption required for OLS to be consistent. Including education in the model therefore introduces an endogeneity bias, and then, the OLS estimator of the education coefficient does not converge to the true causal effect. 

Furthermore, origin and education are correlated. Individuals from certain origin groups have on average lower educational attainment, due to inequalities in access to education. Then, part of the origin on wages operates through the education channel (origin affects education, which in turn affects wages). 
By comparing the short and long regression, including or not the education, we have : 

- when education is omitted in the regression, the origin coefficient captures both the direct effect of origin and its indirect effect through education.
- when education is included in the regression, the origin coefficient is interpreted as the effect of origin conditional on education level. But, as education is endogeneous, the conditional effects is itself biased. In addition, by conditioning on education, we are absorbing part of the total effect of origin on wages : the estimated origin coefficient therefore measures only the direct effect of origin on wages, net of education, and underestimates the total wage inequality due to origin. 

By the omitted variable bias formula (Frisch-Waught):

$$\tilde{\beta}_{origine} = \beta_{origine} + \beta_{educ} \cdot \delta$$

where $\tilde{\beta}_{origine}$ is the coefficient of origin in the short regression of wages on origin ; $\beta_{origine}$ is the coefficient of origin in the long regression of wages on origin and education ; and $\delta = \frac{Cov(educ, origine)}{Var(origine)}$ is the coefficient of the regression of education on origin. As $\beta_{educ} > 0$ and $\delta < 0$ for disadvantaged groups, the bias is negative : we over-estimate in absolute value the wage's disadvantage of theses groups when we omit education. In this case, we attribute the disadvantage for the origins, where it is partly because theses individuals have less education levels, in mean. When we control by education, disadvantage due to origin decreases. 

*Question 2b*

We can choose the variables csp_pere and csp_mere (socio-professional category of the father and mother) as instrumental variables for education. We the construct four binary instruments by regrouping the categories along the PCS nomenclature : 

- father_manager : the father is a senior manager or professional (codes 31 (liberal professions), 33 (senior civil servants), 34 (professors and scientific professions), 35 (information, arts, and entertainment professions), 37 (senior administrative and commercial managers in companies), 38 (engineers and senior technical managers in companies))
- mother_manager : the mother is a senior manager or professional (same codes)
- father_worker : the father is a manual worker (codes 62 (skilled industrial workers), 63 (skilled craft workers), 64 (drivers), 65 (skilled handling, ware-housing, and transport workers), 67 (unskilled industrial workers), 68 (unskilled craft workers), 69 (agricultural workers))
- mother_worker : the mother is a manual worker (same codes)

```{r}
manager   <- c(31, 33, 34, 35, 37, 38)
worker <- c(62, 63, 64, 65, 67, 68, 69)

lastq_emp <- lastq_emp %>%
  mutate(
    csp_pere_num = suppressWarnings(as.numeric(as.character(csp_pere))), # the values are converted to characters and then to numbers
    csp_mere_num = suppressWarnings(as.numeric(as.character(csp_mere))),
    father_manager   = as.numeric(csp_pere_num %in% manager),
    mother_manager   = as.numeric(csp_mere_num %in% manager),
    father_worker = as.numeric(csp_pere_num %in% worker),
    mother_worker = as.numeric(csp_mere_num %in% worker)
  ) %>%
  filter(!is.na(father_manager), !is.na(mother_manager)) # We only keep the rows where "father_manager" is not missing and "mother_manager" is not missing.
# other socio-occupational categories are the default category

# We convert education variable into numeric variable (education_num) to make the 2SLS (education_num is then continuous and we have just one endogeneous variable to instrument, that is possible with 4 instruments)
lastq_emp <- lastq_emp %>%
  mutate(
    education_num = case_when(
      education == ">Bac+2"  ~ 1,
      education == "Bac+2"   ~ 2,
      education == "Bac"     ~ 3,
      education == "CAP-BEP" ~ 4,
      education == "Brevet"  ~ 5,
      education == "Aucun"   ~ 6,
      TRUE                   ~ NA_real_
    )
  )
```
  

Discussion of the validity of the instruments : 2 conditions for an instrumental variable : 

1) Relevance condition (i.e. cov(instrument, education) not equal to 0)
In the sociological and economic literature, we know that parental socio-professional category is highly correlated with the educational attainment of children. Children of senior managers and professionals have substantially higher probabilities of attaining higher education, while children of manual workers have substantially lower probabilities. 
Emprically, we will confirm that by the first-stage F-statistic

2) Exogeneity condition (i.e. cov (instrument, residuals) = 0)
This condition requires that parental ctageory socio-professionnal affects the individual's wage only through the education channel, and not directly.

The respect of this condition is more debatable, because it remains social capital transmission : for example, children of senior managers may benefit from parental networks, job referrals, or behavioural norms that directly improve their labour market outcomes independently of the diploma they hold. In that case, the instrument is invalid.

To identify a model with instrumental variables, the number of instruments must be superior or equal to the number of endogeneous variables. Here, we have only one variable endogeneous (education), and 4 instruments. That improves first-stage precision and allows us to test the joint validity of the instruments via the Sargan-Hansen test. The null hypothesis of this test is that all instruments are valid. 

*Question 2c*

First stage
```{r}
first_stage <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif +
    father_manager + mother_manager + 
    father_worker  + mother_worker,
  se_type = "stata",
  data = lastq_emp
)
summary(first_stage)
```

We examine the coefficients of the four instruments individually in the first-stage regression.
All four coefficients are statistically significant at the 1% level : father_manager (−0.510, p < 0.001), mother_manager (−0.382, p < 0.001), father_worker (+0.233, p < 0.001) and mother_worker (+0.179, p < 0.01).

The signs are consistent with economic intuition : having a parent who is a senior manager or professional is associated with a higher level of education (lower education_num); while having a parent who is a worker is associated with a lower level of education (higher education_num).

To confirm this formally, we conduct a joint significance test of the four instruments.
```{r}
# We use the Wald test with : 
# H0 : the coefficients for all four instruments are jointly zero (irrelevant instruments)
# vs H1 : at least one instrument is relevant
# Simplified model (without instruments)
first_stage_restricted <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif,
  se_type = "stata",
  data = lastq_emp
)

waldtest(first_stage, first_stage_restricted)

F_stat <- 201.3 / 4  # Chi² divided by the number of restrictions
cat("F-stat :", round(F_stat, 2), "\n")

```

The Wald test yields Chi²(4) = 201.3 (p < 0.001), which is equivalent to F ≈ 50.33, far above the conventional threshold of 10. We therefore reject the null hypothesis that the instruments are jointly irrelevant, which confirms that parental socio-professional category is strongly correlated with the individual's education level.


Now, we make the 2SLS :
```{r}
iv_model <- iv_robust(
  logsalhoraire ~ homme + age + age2 + origine + csp_actif + education_num |
                  homme + age + age2 + origine + csp_actif +
                  father_manager + mother_manager +
                  father_worker  + mother_worker,
  se_type = "stata",
  data = lastq_emp
)
summary(iv_model, diagnostics = TRUE)
```
The coefficient of education_num is significative at 1% level and negative, which is consistent with the expected relationship between education and wages (as the level of education increases, so does the wage).

We will compare the OLS estimator of the coefficient of education_num in the OLS regression m6 ; and the 2SLS coefficient of the last regression. 
```{r}
# m6 0LS regression with education_num 
m6_num <- lm_robust(
  logsalhoraire ~ origine + education_num + homme + age + age2 + csp_actif,
  se_type = "stata",
  data = lastq_emp
)
summary(m6_num)

# comparison between OLS and 2SLS
library(texreg)
screenreg(
  list(m6_num, iv_model),
  custom.model.names = c("OLS", "2SLS"),
  digits = 4
)
```
*Question 2d*

By comparing the OLS and 2SLS estimates, we can assess the impact of correcting for the endogeneity of education. The coefficient on education_num is larger in absolute value in the 2SLS specification (−0.088) than in OLS (−0.052), suggesting that OLS underestimated the causal return to education. Based on the 2SLS regression, all other things being equal, moving down one educational level reduces earnings by around 8.8%. 

Regarding origin, the coefficients are stable between OLS and 2SLS. No origin group changes significance category substantially. Theses results confirm that the endogeneity bias of education does not distort the estimated origin wage gaps. 
Southern European origin remains positively significant in both specifications (+5.2% in OLS, +6.6% in 2SLS). Based on the 2SLS regression, all other things being equal (in terms of education, gender, age and socio-professional category), a person of Southern European origin earns on average 6.6% more than a person of French origin.

For other origin groups, no significant negative wage penalty is detected once education and occupational category are controlled for.
Then, adjusting for the endogeneity of education does not substantially alter our conclusions regarding wage inequalities linked to origins.

However, the exogeneity condition for the instruments (parents’ socio-economic status) remains impossible to verify directly. If parents’ socio-economic status transmits social capital beyond education, the 2SLS estimates remain biased. 

# version 2 : with the instrument as a continuous variable

```{r}
csp_continuous <- function(x) {
  case_when(
    x %in% c(31,33,34,35,37,38, 74)          ~ 5,  # cadres + anciens cadres
    x %in% c(21,22,23, 72)                    ~ 4,  # artisans/chefs d'entreprise
    x %in% c(42,43,44,45,46,47,48, 75)        ~ 3,  # professions intermédiaires
    x %in% c(52,53,54,55,56, 77)              ~ 2,  # employés
    x %in% c(10,11,12,13, 71)                 ~ 2,  # agriculteurs
    x %in% c(62,63,64,65,67,68,69, 78)        ~ 1,  # ouvriers + anciens ouvriers
    x %in% c(81,82)                           ~ 1,  # sans activité
    TRUE                                      ~ NA_real_
  )
}

lastq_emp <- lastq_emp %>%
  mutate(
    csp_pere_num  = suppressWarnings(as.numeric(as.character(csp_pere))),
    csp_mere_num  = suppressWarnings(as.numeric(as.character(csp_mere))),
    level_pere    = csp_continuous(csp_pere_num),
    level_mere    = csp_continuous(csp_mere_num)
  )
```

```{r}
# First stage
first_stage_v2 <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif +
    level_pere + level_mere,
  se_type = "stata",
  data = lastq_emp
)
summary(first_stage_v2)

# 2SLS
iv_model_v2 <- iv_robust(
  logsalhoraire ~ homme + age + age2 + origine + csp_actif + education_num |
                  homme + age + age2 + origine + csp_actif +
                  level_pere + level_mere,
  se_type = "stata",
  data = lastq_emp
)
summary(iv_model_v2, diagnostics = TRUE)
```
Pb : there are only 2 291 observations against 4 046 in the previous regression. The sub-sample for which both parents' CSP levels are observed may not be representative of the full employed population, limiting the external validity of this specification

```{r}
# Combien ont level_pere manquant ?
cat("NA level_pere :", sum(is.na(lastq_emp$level_pere)), "\n")
cat("NA level_mere :", sum(is.na(lastq_emp$level_mere)), "\n")

# Quelles CSP du père produisent des NA ?
lastq_emp %>%
  mutate(na_level = is.na(level_pere)) %>%
  count(csp_pere, na_level) %>%
  filter(na_level == TRUE) %>%
  arrange(desc(n))

# Distribution des CSP manquantes
lastq_emp %>%
  mutate(
    csp_pere_num = suppressWarnings(as.numeric(as.character(csp_pere))),
    dans_grille = case_when(
      csp_pere_num %in% c(31,33,34,35,37,38,74) ~ "cadres",
      csp_pere_num %in% c(21,22,23,72)           ~ "artisans",
      csp_pere_num %in% c(42,43,44,45,46,47,48,75) ~ "intermediaires",
      csp_pere_num %in% c(52,53,54,55,56,77)     ~ "employes",
      csp_pere_num %in% c(10,11,12,13,71)        ~ "agriculteurs",
      csp_pere_num %in% c(62,63,64,65,67,68,69,78) ~ "ouvriers",
      csp_pere_num %in% c(81,82)                 ~ "sans_activite",
      TRUE                                       ~ "non_classe"
    )
  ) %>%
  count(dans_grille) %>%
  mutate(pct = round(100*n/sum(n), 1)) %>%
  arrange(desc(n))
  
```  
```{r}
#lastq_emp %>%
  mutate(
    csp_mere_num = suppressWarnings(as.numeric(as.character(csp_mere)))
  ) %>%
  count(csp_mere, sort = TRUE) %>%
  head(20)
```
A lot of mother have "00" as a category. So we decide to take only the CSP level of the father as the instrument.
```{r}

# Creation of a sub-sample without NA on level_pere
lastq_emp_v3 <- lastq_emp %>%
  filter(!is.na(level_pere), !is.na(education_num))

cat("N échantillon v3 :", nrow(lastq_emp_v3), "\n")

# First stage
first_stage_v3 <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif + level_pere,
  se_type = "stata",
  data = lastq_emp_v3
)
summary(first_stage_v3)

# Restricted model on the same sample
first_stage_v3_restricted <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif,
  se_type = "stata",
  data = lastq_emp_v3  # <-- même données
)

# test
waldtest(first_stage_v3, first_stage_v3_restricted)

# 2SLS 
iv_model_v3 <- iv_robust(
  logsalhoraire ~ homme + age + age2 + origine + csp_actif + education_num |
                  homme + age + age2 + origine + csp_actif + level_pere,
  se_type = "stata",
  data = lastq_emp_v3  # <-- même données
)
summary(iv_model_v3)

# Comparison between regressions
screenreg(
  list(m6_num, iv_model, iv_model_v3),
  custom.model.names = c("MCO", "2SLS v1 (binaire)", "2SLS v3 (ordinal père)"),
  digits = 4,
  suppress.warnings = TRUE
)
```
!! There are only 3 700 observations compared with 4 046 for the MCO and 2SLS v1. The difference is due to the 346 fathers with an unknown socio-professional category (csp_pere = ‘00’).

Correcting for the endogeneity of education increases the estimated return to education in absolute value (from −0.052 in OLS to −0.088 in 2SLS v1 and −0.076 in 2SLS v3), confirming an upward bias in OLS estimates. 

The test : We reject H0 (zero coefficient for level_pere) by a significant margin. The relevance condition is also largely satisfied for 2SLS v3. level_pere alone is a stronger instrument than the four binary instruments combined (its individual F-statistic is three times higher). This is because the ordinal variable captures the continuous variation in parental socio-economic status more effectively, whereas the binary variables captured only the extremes (managers vs workers).


Then, the 2SLS v1 model suggests that a one-level increase in educational attainment (for example, moving from a Baccalaureate to a CAP-BEP in our coding) increases wages by approximately 8.8%, compared with 5.2% in the OLS model; whereas the 2SLS v3 model estimates this effect at 7.6%. The difference between the two 2SLS models is moderate and can be explained by the difference in the sample and the nature of the instruments.

No origin group displays a statistically significant negative wage gap once education and occupational category are controlled for, in any of the three models. The only positive effect (Southern European origin) is significant in OLS (+5.2%) and 2SLS v1 (+6.6%) but not in 2SLS v3, suggesting this result lacks full robustness across instrument specifications.  


# Version 3 : education as a continuous variable


We treat education as a continuous variable (education_num, coded from 1 for the highest level to 6 for no diploma), which allows us to instrument a single endogenous variable rather than five dummies simultaneously.

!! On recrée le dataset avec les mères avec NA car on les a enlevées dans la précédente version 
```{r}

lastq_emp <- df %>%
  filter(t == 6, acteu == "Employed", !is.na(salhoraire))

cat("N de départ :", nrow(lastq_emp), "\n")
# Doit donner 4046

# Recréer toutes les variables nécessaires
lastq_emp <- lastq_emp %>%
  mutate(
    education_num = case_when(
      education == ">Bac+2"  ~ 1,
      education == "Bac+2"   ~ 2,
      education == "Bac"     ~ 3,
      education == "CAP-BEP" ~ 4,
      education == "Brevet"  ~ 5,
      education == "Aucun"   ~ 6,
      TRUE ~ NA_real_
    ),
    csp_pere_num   = suppressWarnings(as.numeric(as.character(csp_pere))),
    csp_mere_num   = suppressWarnings(as.numeric(as.character(csp_mere))),
    level_pere     = csp_to_level(csp_pere_num),
    level_mere     = csp_to_level(csp_mere_num),
    level_mere_imp = ifelse(is.na(level_mere), 0, level_mere)
  )

cat("NA level_pere :", sum(is.na(lastq_emp$level_pere)), "\n")
cat("NA level_mere_imp :", sum(is.na(lastq_emp$level_mere_imp)), "\n")

# Filtre final : uniquement level_pere et education_num
lastq_emp <- lastq_emp %>%
  filter(!is.na(level_pere), !is.na(education_num))

cat("N final :", nrow(lastq_emp), "\n")

```

```{r}

csp_to_level <- function(x) {
  case_when(
    # Niveau 6 : Cadres et professions intellectuelles supérieures
    x %in% c(31, 33, 34, 35, 37, 38,  # actifs
              74)                        # anciens cadres
    ~ 6,
    
    # Niveau 5 : Artisans, commerçants, chefs d'entreprise
    x %in% c(21, 22, 23,               # actifs
              72)                        # anciens
    ~ 5,
    
    # Niveau 4 : Professions intermédiaires
    x %in% c(42, 43, 44, 45, 46, 47, 48,  # actifs
              75)                            # anciens
    ~ 4,
    
    # Niveau 3 : Employés
    x %in% c(52, 53, 54, 55, 56,       # actifs
              77)                        # anciens
    ~ 3,
    
    # Niveau 2 : Agriculteurs exploitants
    x %in% c(10, 11, 12, 13,           # actifs
              71)                        # anciens
    ~ 2,
    
    # Niveau 1 : Ouvriers + sans activité
    x %in% c(62, 63, 64, 65, 67, 68, 69,  # ouvriers actifs
              78,                            # anciens ouvriers
              81, 82)                        # sans activité
    ~ 1,
    
    # NA : non renseigné (code 00) ou valeur manquante
    x == 0 | is.na(x) ~ NA_real_
  )
}

# Application aux deux parents
lastq_emp <- lastq_emp %>%
  mutate(
    csp_pere_num = suppressWarnings(as.numeric(as.character(csp_pere))),
    csp_mere_num = suppressWarnings(as.numeric(as.character(csp_mere))),
    level_pere   = csp_to_level(csp_pere_num),
    level_mere   = csp_to_level(csp_mere_num)
  )

# Vérification
cat("NA level_pere :", sum(is.na(lastq_emp$level_pere)), "\n")
cat("NA level_mere :", sum(is.na(lastq_emp$level_mere)), "\n")

# Distribution des niveaux
table(lastq_emp$level_pere, useNA = "always")
table(lastq_emp$level_mere, useNA = "always")


```

The construction of level_pere and level_mere produces missing values only for individuals whose parent's socio-professional category is coded "00" (not stated). For fathers, this concerns 346 individuals (8.6% of the sample). For mothers, however, 1,561 individuals (38.6%) have a missing maternal CSP — a far larger share that would severely reduce the sample if treated as missing.

We therefore impute level_mere = 0 when csp_mere = "00". We consider that a child raised by a mother with no professional activity is likely to have lower educational attainment than one raised by a mother with even a modest occupational status, all else equal. 

This imputation allows us to retain the full sample of 4 046 observations and to use both level_pere and level_mere as instruments simultaneously, yielding an overidentified model and enabling a Sargan test of instrument validity.
```{r}
lastq_emp <- lastq_emp %>%
  mutate(level_mere_imp = ifelse(is.na(level_mere), 0, level_mere))

# Vérification
cat("NA level_mere_imp :", sum(is.na(lastq_emp$level_mere_imp)), "\n")
table(lastq_emp$level_mere_imp, useNA = "always")
```
```{r}
lastq_emp <- lastq_emp %>%
  filter(!is.na(level_pere), !is.na(education_num))

cat("N :", nrow(lastq_emp), "\n")
```
# First stage
```{r}
first_stage <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif +
    level_pere + level_mere_imp,
  se_type = "stata",
  data = lastq_emp
)
summary(first_stage)
```
The relevance condition requires that the parental CSP level variables are correlated with the individual's education level, conditional on the other covariates. This condition is assessed through the first-stage regression of education_num on level_pere, level_mere_imp, and all exogenous controls. Both instruments are statistically significant at the 1% level : the coefficient on level_pere is −0.124 (p < 0.001) and the coefficient on level_mere_imp is −0.068 (p < 0.001). The signs are consistent with economic intuition : a higher parental occupational level (coded from 1 (manual workers) to 6 (senior managers)), is associated with a lower value of education_num, which corresponds to a higher level of educational attainment since education_num is coded in decreasing order of qualification. Then, children of higher-status parents systematically attain higher levels of education, all else equal.


```{r}
# relevance test
first_stage_restricted <- lm_robust(
  education_num ~ homme + age + age2 + origine + csp_actif,
  se_type = "stata",
  data = lastq_emp
)
waldtest(first_stage, first_stage_restricted)
```


```{r}
# F-stat 
r2_full <- summary(first_stage)$r.squared
r2_res  <- summary(first_stage_restricted)$r.squared
n <- nrow(lastq_emp)
k <- length(coef(first_stage))
q <- 2  # nombre d'instruments
F_stat <- ((r2_full - r2_res) / q) / ((1 - r2_full) / (n - k))
cat("F-stat instruments :", round(F_stat, 2), "\n")
```
A Wald test comparing the full first-stage model to a restricted model excluding both instruments yields Chi²(2) = 192.6 (p < 0.001), equivalent to an F-statistic of 95.94, far above the conventional threshold of 10. The relevance condition is therefore strongly satisfied.
```{r}
# 2SLS
iv_model <- iv_robust(
  logsalhoraire ~ homme + age + age2 + origine + csp_actif + education_num |
                  homme + age + age2 + origine + csp_actif +
                  level_pere + level_mere_imp,
  se_type = "stata",
  data = lastq_emp
)
summary(iv_model, diagnostics = TRUE)
```
```{r}
# Comparative board
m6_num_iv <- lm_robust(
  logsalhoraire ~ homme + age + age2 + origine + csp_actif + education_num,
  se_type = "stata",
  data = lastq_emp
)

screenreg(
  list(m6_num_iv, iv_model),
  custom.model.names = c("MCO", "2SLS (level_pere + level_mere_imp)"),
  digits = 4,
  suppress.warnings = TRUE
)
```
The comparison between OLS and 2SLS :
Correcting for the endogeneity of education increases the estimated return to education in absolute value, from −0.052 in OLS to −0.077 in 2SLS, suggesting a downward bias in OLS estimates consistent with the endogeneity argument (q° 2a).

Regarding origin, the results are stable : no origin group displays a statistically significant wage gap in either specification. So the endogeneity bias of education does not distort our conclusions about wage inequalities by origin, no evidence of direct wage discrimination at equal post and qualification is detected.


*Question 3*
*Question 3a*

We estimate a pooled OLS model on the panel sub-sample of employed individuals observed at t=1 and t=6, with clustered standard errors at the individual level to account for serial correlation within individuals. We make the random effects assumption that individual effects αᵢ are uncorrelated with the regressors.

```{r}
# We create the panel subset for active individuals for t=1 and t=6 

panel_q3 <- df %>%
  filter(t %in% c(1, 6),
         acteu == "Employed",
         !is.na(salhoraire)) %>%
  mutate(
    # We create temporal dummies : 1 if t=6, 0 if t=1

    d_last = as.numeric(t == 6),
    
    age2 = age^2
  )

cat("N observations :", nrow(panel_q3), "\n")
cat("N individuals :", n_distinct(panel_q3$id_individu), "\n")


# Fit the pooled OLS model
# We create a temporal dummy d_last to catch the common effect of the period
pooled_ols <- lm(logsalhoraire ~ origine + education + homme + age + age2 + csp_actif + d_last, data = panel_q3)

# Clustered standard errors at the "id" level
# (comme vcovCL(pooled_ols, cluster = ~id) dans le TD)
clustered_se <- vcovCL(pooled_ols, cluster = ~id_individu)

# Summary of the model with clustered standard errors
coeftest(pooled_ols, vcov = clustered_se)
```
d_last coefficient is +0.015 (p = 0.003). This means that, between the first and last waves observed, wages increased by an average of 1.5% all other things being equal. This captures the conjunctural effect common to all individuals.

 Education remains the strongest determinant of wages, with a significant gradient across levels. 
 Among origin groups, only Southern European origin shows a significant positive coefficient (+4.5%, p < 0.05)
 
Finally, the results are consistent with the cross-sectional estimates of Question 1.

*Question 3b*

We can include the origine variable in the pooled OLS model estimated in the previous question because this estimator does not eliminate time-invariant regressors (as the origin).

However, this model does not resolve potential endogeneity problems because it relies on an unrealistic assumption. Indeed, in the pooled OLS, we consider that the individual effects αi are uncorrelated with all regressors, including origine. But this assumption is very strong and unlikely to hold in practice. In fact, unobserved individual characteristics such as family networks or cultural capital are potentially correlated with origin. It violates the exogeneity condition. If this assumption fails, the pooled OLS estimator is then inconsistent.

Furthermore, we can't use the within estimator to solve the endogeneity problem because the variable origin is time-invariant. Indeed, the origin variable does not change over time for a given individual, the parents’ nationality is fixed at birth. So with the within estimator, the coefficient of origin would not be identified. 

# 2. Extensive Margin (Employed vs. Unemployed)

*Question 4a* :

The phrasing "exogeneity of residuals and individual fixed effects is assumed" points to the Random Effects (RE) framework. In this model, we assume that the unobserved individual heterogeneity ($\alpha_i$) is uncorrelated with the explanatory variables ($E(\alpha_i | X_{it}) = 0$). To estimate this, we first need to build a panel data frame containing all active individuals across all available periods ($t=1$ to $6$), not just the last quarter.


The choice to employ a Linear Probability Model (LPM) rather than a non-linear specification, such as a Probit or Logit model, is driven by both interpretability and methodological flexibility. The primary advantage of the LPM is that its estimated coefficients directly represent the average marginal effect of the regressors on the probability of unemployment, allowing for a straightforward and intuitive interpretation in percentage points. Furthermore, non-linear models face significant structural limitations in panel data settings, most notably the incidental parameters problem and severe complications when integrating instrumental variables (IV) for dynamic models, which are required in subsequent steps of this analysis. Although the LPM inherently produces heteroskedasticity and can theoretically yield predicted probabilities outside the logical [0, 1] bounds, the former issue is effectively addressed by computing clustered robust standard errors. Given that the main objective is to estimate average partial effects rather than predicting individual probabilities, the LPM remains a robust, transparent, and widely supported framework in applied microeconometrics.

```{r}
# We create the panel subset for active individuals across all periods
panel_active <- df %>%
  dplyr::filter(acteu != "Inactive") %>%
  mutate(
    unemp = ifelse(acteu == "Unemployed", 1, 0),
    csp_actif_clean = droplevels(csp_actif)
  )

cat("N =", nrow(panel_active), "active individuals*quarters \n")

#We declare the dataset as a panel using pdata.frame
p_active <- pdata.frame(panel_active, index = c("id_individu", "t"))

#We estimate the Linear Probability Model with Random Effects
lpm_re <- plm(unemp ~ origine_cat + homme + age + age2 + csp_actif_clean, 
              data = p_active, 
              model = "random")

summary(lpm_re)

# We compute clustered standard errors at the individual level to account for serial correlation
coeftest(lpm_re, vcov = vcovHC(lpm_re, type = "HC1", cluster = "group"))
```

The analysis employs a Linear Probability Model (LPM) with Random Effects to estimate the determinants of unemployment. Because the dependent variable is binary, taking the value of 1 for unemployed individuals and 0 for active employed individuals, the estimated coefficients can be directly interpreted as the change in the probability of being unemployed, expressed in percentage points, ceteris paribus. The Random Effects specification relies on the assumption that unobserved individual heterogeneity—such as innate characteristics that remain constant over time—is strictly uncorrelated with the explanatory variables in the model. Furthermore, since the LPM intrinsically generates heteroskedasticity and the panel structure implies serial correlation among observations of the same individual, robust standard errors clustered at the individual level are utilized to ensure valid statistical inference.

Regarding the primary variable of interest, the origin of the individuals, the reference category is set to individuals of French origin. The results reveal a significant labor market penalty for several groups. Holding age, gender, and socio-professional category constant, individuals of Maghrebin origin face a probability of being unemployed that is 10.87 percentage points higher than their French-origin counterparts, an effect that is highly statistically significant. Similarly, individuals originating from the Rest of Africa and Southern Europe exhibit higher probabilities of being unemployed, by 8.01 and 3.75 percentage points respectively, both significant at the 5% level. Conversely, the coefficients for other origins, namely Eastern Europe, Northern Europe, the Middle East, and the Rest of the World, do not show statistically significant differences from the reference group.

The control variables align with standard labor market expectations. Gender plays a significant role, with men exhibiting a 2.28 percentage point lower probability of being unemployed compared to women. The effect of age is captured through both linear and quadratic terms, revealing a U-shaped relationship where the probability of unemployment initially decreases with age before slightly increasing for older individuals. Furthermore, socio-professional categories strongly correlate with employment status, most notably demonstrated by the extreme mechanical correlation for the "Unemployed never worked" category, which shows an 83.08 percentage point increase in unemployment probability relative to the omitted baseline category.

Despite these findings, the core assumption of the Random Effects model—that unobserved individual effects are uncorrelated with the regressors—remains a strong and potentially restrictive premise. It is highly plausible that unobserved traits, such as personal motivation or social networks, are correlated with an individual's origin or socio-professional category, thereby introducing an omitted variable bias into the current estimates. While a Fixed Effects (Within) model would traditionally correct for this endogeneity by eliminating time-invariant unobserved heterogeneity, implementing it in this context poses a fundamental structural issue: the Within transformation would simultaneously eliminate all time-invariant explanatory variables, thereby making it impossible to identify the causal effect of origin.


*Question 4b* : 

```{r}
# --- Question 4b ---

# We estimate the probability linear model with fixed effects (Within)
lpm_fe <- plm(unemp ~ origine_cat + homme + age + age2 + csp_actif_clean, 
              data = p_active, 
              model = "within")

# Results
summary(lpm_fe)
```
To address the potential omitted variable bias inherent in the Random Effects specification, a Fixed Effects (Within) model is estimated. This approach rigorously controls for all unobserved, time-invariant individual heterogeneity by allowing it to be arbitrarily correlated with the explanatory variables. However, examining the estimation results reveals a fundamental structural limitation for this specific research objective: the variables corresponding to the individual's origin (origine_cat) and gender (homme) are completely omitted from the coefficient table.

This omission is a direct consequence of the mathematical mechanics underlying the Within transformation. To eliminate the unobserved individual fixed effects, the estimator subtracts the individual-specific time average from each observation. Because demographic characteristics such as origin and gender are strictly time-invariant for a given individual across the observed quarters, their value at any given time t is exactly equal to their individual time average ($X_{it} = \bar{X}_i$). Consequently, the within-group differencing perfectly cancels them out ($X_{it} - \bar{X}_i = 0$), rendering the estimation of their coefficients impossible due to perfect collinearity with the fixed effects.

Therefore, while the Fixed Effects model is theoretically more robust to endogeneity stemming from unobserved individual traits, it completely precludes the identification and estimation of the causal effect of origin on the probability of being unemployed. This structural constraint ultimately justifies the reliance on the Random Effects model to quantify the labor market penalty associated with an individual's origin, albeit with the necessary caveat that those estimates might still capture other unobserved, correlated factors.

*Question 5a* :

```{r}
# We create the lagged variables and differences taking the panel structure into account
p_active_dyn <- p_active %>%
  group_by(id_individu) %>%
  arrange(t) %>%
  mutate(
    # Lagged levels
    l.unemp = dplyr::lag(unemp, 1),
    l2.unemp = dplyr::lag(unemp, 2),
    
    # First differences
    d.unemp = unemp - l.unemp,
    d.l.unemp = l.unemp - l2.unemp # This corresponds to \Delta Y_{t-1}
  ) %>%
  ungroup()

# We estimate the First-Differenced equation using IV (Anderson-Hsiao approach)
# We instrument the differenced lag (d.l.unemp) with the twice-lagged level (l2.unemp)
dyn_iv <- ivreg(d.unemp ~ d.l.unemp | l2.unemp, data = p_active_dyn)

summary(dyn_iv)
coeftest(dyn_iv, vcov = vcovCL, cluster = ~id_individu)
```
Note that the control variables present in the previous models (origin, gender, age) are intentionally omitted from this dynamic specification. Because we apply a First-Difference transformation to eliminate the individual fixed effect $\alpha_i$, all time-invariant variables (such as origine_cat and homme) mathematically disappear. Furthermore, since age increases by a constant exact amount for everyone each quarter, its first difference is a constant which is simply absorbed by the model's intercept. Therefore, the equation is solely focused on the dynamics of the dependent variable.

*Questions 5b* :

The estimation results from the Anderson-Hsiao instrumental variables approach yield a highly significant coefficient of approximately 0.364 for the first-differenced lagged dependent variable (d.l.unemp). Importantly, this statistical significance holds even after applying robust standard errors clustered at the individual level, which correctly adjusts our inference for both arbitrary heteroskedasticity and within-individual serial correlation.

Economically, this coefficient quantifies the magnitude of "true state dependence" in unemployment, a phenomenon frequently referred to in labor economics as the "scarring effect." Because the dependent variable is binary (estimated via a Linear Probability Model framework), the coefficient can be interpreted directly as a marginal effect on probability. Specifically, it indicates that the mere fact of having been unemployed in the previous quarter directly increases the probability of remaining unemployed in the current quarter by 36.4 percentage points.

Crucially, the methodological design of this dynamic model allows for a strong causal interpretation. By applying the First-Difference transformation, we have completely purged the model of time-invariant unobserved individual heterogeneity—such as innate ability, permanent motivation, or structural disadvantages related to an individual's origin. Therefore, this 36.4 percentage point penalty does not merely reflect "spurious state dependence," a scenario where certain individuals simply possess a higher baseline propensity for unemployment due to unobserved traits. Instead, it demonstrates an endogenous mechanism: the unemployment spell itself actively damages an individual's future employment prospects, likely through the depreciation of human capital or by acting as a negative signal to prospective employers.

*Question 6a* : 

Afin de considérer un modèle avec un effet fixe, il est nécessaire de l'adpater au cas de la variable binaire. Suppose there exists a continuous variable $$Y^*$$, the latent variable, such that $$Y=\mathbf{1}\{Y^* \geq 0\}$$ and suppose that $$Y^*$$ is generated by a linear model with a fixed effect : 

$$
Y_{it} = X_{it}'\beta_0 + \alpha_i + \varepsilon_{it}
$$
where $-\varepsilon$ is independent of $X$ and has distribution function $F$. Then
$$
E(Y_{it} \mid X_{it}, \alpha_i) \\
= P(Y_{it}=1 \mid X_{it}, \alpha_i) \\
= P(X_{it}\beta + \alpha_i + \varepsilon_{it} \geq 0 \mid X_{it}, \alpha_i) \\
= P(-\varepsilon_{it} \leq X_{it}\beta + \alpha_i \mid X_{it}, \alpha_i) \\
 \\
= F(X_{it}' \beta + \alpha_i)
$$
If we try to apply the first difference method, we obtain : 
$$
E(Y_{it} - Y_{i,t-1} \mid X_{it}, X_{i,t-1}, \alpha_i)  \\
= E(Y_{it} \mid X_{it}, X_{i,t-1}, \alpha_i) - E(Y_{i,t-1} \mid X_{it}, X_{i,t-1}, \alpha_i) \\
= F(X_{it}' \beta + \alpha_i) - F(X_{i,t-1}' \beta + \alpha_i)
$$
Yet, the function F is not linear, as it is a distribution function of a random normal variable or the logistic distribution function. This do not allow us to eliminate the fixed effect as the linear model, in order to estimate $\beta$ without biais. In conclusion, we cannot use the first difference panel model. 

If we try to apply the within panel method, we obtain : Q° on dvp le fait que parfois c négatif / positif ?? 
$$
E\big(Y_{it} - \bar{Y}_i \mid X_{it}, \alpha_i \big)\\ 
= E\big(Y_{it} \mid X_{it}, \alpha_i \big) - E\big(\bar{Y}_i \mid X_{it}, \alpha_i \big) \\
= E\big(Y_{it} \mid X_{it}, \alpha_i \big) - E(\frac{1}{T} \sum_{t=1}^{T}{Y}_{it} \mid X_{it}, \alpha_i \big) \\
= E\big(Y_{it} \mid X_{it}, \alpha_i \big) - \frac{1}{T}\sum_{t=1}^{T}E({Y}_{it} \mid X_{it}, \alpha_i \big) \\
= F(X_{it}' \beta + \alpha_i) - \frac{1}{T} \sum_{t=1}^{T} F(X_{it}' \beta + \alpha_i)
$$
Yet, the function F is not linear, as it is a distribution function of a random normal variable or the logistic distribution function. This do not allow us to eliminate the fixed effect as the linear model, in order to estimate $\beta$ without biais. In conclusion, we cannot use the within panel model. 


AU BROUILLON AU CAS OU les formules de logit et probit : 

So we have : 
$$
E(Y \mid X) = F(X' \beta_0)
$$
$$
P(Y_{it}=1 \mid X_{it}, \alpha_i) = \frac{\exp(X_{it}\beta + \alpha_i)}{1 + \exp(X_{it}\beta + \alpha_i)}
$$
$$
P(Y_{it}=1 \mid X_{it}, \alpha_i) = \Phi(X_{it}\beta + \alpha_i)
$$
$$
Y_{it} - Y_{i,t-1} = (X_{it} - X_{i,t-1})\beta + (\varepsilon_{it} - \varepsilon_{i,t-1})
$$

*Question 6b* : 

First, we create the correct subset to restrict our analysis to cross-sectional data using the last quarter and focusing on active individuals (employed or unemployed). We create then the indicator of being unemployed. 
```{r}
# subset 
lastq <- df %>%
  dplyr::filter(t == 6, acteu != "Inactive") 
cat("N =", nrow(lastq), "active individuals")

# recoding for modeling 
lastq <- lastq %>% mutate(unemp = ifelse(lastq$acteu == "Unemployed", 1, 0))

# Checking
freq(lastq$unemp)
freq(lastq$acteu)
```
Our analysis will hold on 5030, including 324 individuals who are unemployed. 

We suppose that the variable csp_actif in the data followed the french INSEE convention : for the undemployed individuals, the csp_actif indicates the CSP associated with their last employment. 
We have to elminate the 
```{r}
freq(lastq$csp_actif) # modality empty ("") without any individualis in it 
lastq$csp_actif_clean <- droplevels(lastq$csp_actif)
levels(lastq$csp_actif_clean) 

# check 
table(lastq$csp_actif,lastq$csp_actif_clean) # only terms in the diagonal : OK 
```


We begin by estimating a probit model. 
```{r}
probit_model <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "probit"), data = lastq)
summary(probit_model) # we interpret the sign and the significance of the coefficients, but not their values 
levels(lastq$origine_cat)
```
AIC: 2208.2
Number of Fisher Scoring iterations: 15

We compute the estimated Average Marginal Effect (AME) in the probit model. 
```{r}
AME_probit <- margins(probit_model)
summary(AME_probit)
```
The variable origin is a categorical variable. Formally, when we add this variable in the model, we include all the indicators for the categories, except one that is the category of reference (France in this case). Analyzing the average marginal effect of the variable origin means to look at the AME of each dummy variable, in reference to the French origin. 

Indeed, if we denote \(X_j\), one of the dummies variable origin, the average effect of switching \(X_j\) from 0 to 1 is then:

$$
E\Big[ F(X_{-j}' \beta_{0,-j} + \beta_{0j}) - F(X_{-j}' \beta_{0,-j}) \Big]
$$
In addition, in a binary model, the units of the AME are points of pourcentage. We will denote this unit pp. 

Only one origin has a significant effect at 1% on the probability of beoing unemployed : it is Maghrebian origin. All else being equal, to have at least one parent who hold an origin of the Maghreb rather than having both parents who are French increase of 9,2 pp on average the probability of being unemployed.  
The other variable are not statistically significant. 


We estimate a logit model. 
```{r}
logit_model <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "logit"), data = lastq)
summary(logit_model) # we interpret the sign and the significance of the coefficients, but not their values 
```
AIC: 2372.3
Number of Fisher Scoring iterations: 14

We compute the Average Marginal Effect (AME) in the probit model. 
```{r}
AME_logit <- margins(logit_model)
summary(AME_logit)
```
The result are quite identical with the ones of the probit model. Only the Maghrebian origin is statiscally significant, and it is at 1%, and the estimated AME is a bit more important than the probit model : 9,3 pp rather than 9,2 pp. 

To be able to observe some differences comparing two non-French origins, we test some models behind.  

```{r}
lastq$origine_cat <- relevel(lastq$origine_cat, ref = "Middle East")
logit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "logit"), data = lastq)
summary(margins(logit_model2))
```
No variable has a significant effect on the probability studied. 

```{r}
lastq$origine_cat <- relevel(lastq$origine_cat, ref = "Rest of Africa")
logit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "logit"), data = lastq)
summary(margins(logit_model2))
```
No variable has a significant effect on the probability studied. 

```{r}
lastq$origine_cat <- relevel(lastq$origine_cat, ref = "Maghreb")
logit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2, family = binomial(link = "logit"), data = lastq)
summary(margins(logit_model2))
```
To see some differences, we try to compare the effect of being from another origin than Maghrebian ones, has it was the only one to be significant previously. Except French people, the individuals which have less chance of being unemployed in average, all else being equal, than the individuals with Maghrebian origins are : 
- the one with Northern European origins (-9,1 pp, significant at 5%)
- the ones with Southern European origins (-7,8 pp, significant at 1%)
- the ones with "Rest of the world origin" ("Rest of the world origin" + Southern Asia) (-11,3 pp, significant at 1%). 
The effect of the other origins cannot be stated as statiscly different than the Maghrebian ones. 

This provides a more in-depth analysis of the effect of origins in the labor market.  

*Question 6c* : 

We create the indicator of having a higher education degree (modality "1" of the variable education in the dataset). 
```{r}
lastq <- lastq %>% mutate(highdegree = ifelse(lastq$education == ">Bac+2", 1, 0))
# Checking
#freq(lastq$highdegree) # 975 individuals
#freq(lastq$education)
```

Before using the Rivers-Vuong method, we include the new indicator as an explanatory variable, together with "origin" and the other relevant covariates, using a logit model. 
```{r}
logit_model2 <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2 + highdegree, family = binomial(link = "logit"), data = lastq)
summary(logit_model2) # we interpret the sign and the significance of the coefficients, but not their values 
summary(AME_logit)
```

For the same reason than in question 2, to control for education level may raise an endogeneity problem. In the following treatment, we will ignore the fact that the Rivers-Vuong method do no work for an binary endogeneous variable. We apply this method. Note that we sill use a categorical variable, that correspond to a set of indicators ued as intruments of the endogeneous variable.  

The variable highdegree is endogeneous. We consider this model : 
$$Y=\mathbf{1}\{X'_1\beta_1 + X_2\beta_2+ \epsilon \geq 0\}$$, where is $$X_1$$ the matrix of the control variable and $$X_2$$ is the indicator $$highdegree$$, which is endogeneous, meaning not independant from $$\epsilon$$. 

We choose the same instrument $$Z$$ that before : the csp of the father. It supposed to have a direct effect on $$higherdiplom$$ but not on $$Y$$, meaning that we supposed that : 

$$ 
X_2 = X'_1\gamma_1 + Z\gamma_2 + \eta
$$

where $$\epsilon = \rho\eta + \zeta $$

The model is now : 
$$
Y=\mathbf{1}\{X'_1\beta_1 + X_2\beta_2+ \rho\eta + \zeta \geq 0\}
$$


Firstly, we compute the related OLS regression to obtain en estimation of $$\eta$$. 
```{r}
mod <- lm(higherdegree ~ origine_cat + homme + csp_actif_clean + age + age2 +  csp_pere, data = lastq) # estimation du model 
lastq$etahat<- mod$residuals  # si pb de lignes 
```

SI BESOIN (pas de pb de NA normalement)
```{r}
# On s'assure que le modèle garde la trace des lignes exclues
mod <- lm(highdegree ~ origine_cat + homme + csp_actif_clean + age + age2 + csp_pere, 
          data = lastq, 
          na.action = na.exclude) 

# L'attribution se fera parfaitement, même s'il y avait des NA
lastq$etahat <- residuals(mod)
```


Then, we want to compute the logit model on  $$Y=\mathbf{1}\{X'_1\beta_1 + X_2\beta_2+ \beta_3\widehat\eta + \zeta \geq 0\}$$. 
We construct the variable $$\zeta$$, such as $$\zeta = \epsilon - \rho\hat\eta$$
```{r}

```


```{r}
model_instru <- glm(unemp ~ origine_cat + homme + csp_actif_clean + age + age2 + highdegree + etahat, family = binomial(link = "logit"), data = lastq)
summary(model_instru) 
```


